### 2020-03-29 Paper/Paper.tex
Added folder structure

\documentclass[12pt,letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\begin{document}
\maketitle
\section*{Deliverables}
\begin{enumerate}
\item Load the data
\item Performance benchmarking
\item PCA
\item Thresholding PCA
\item Simple Principal Components
\item CPEV versus Sparsity plots
\item SCoTLASS
\item SPCA
\end{enumerate}
\end{document}

### 2020-03-29 Paper/Proposal.tex
Added folder structure

\documentclass[10pt,letterpaper]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{apacite}
\usepackage{ntheorem}
\newtheorem{hyp}{Hypothesis}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\author{
	Kho, Lee\\
	\textit{ltk224}
	\and
	Peeters, Jonas\\
	\textit{jp5642}
}
\title{DS-GA 1013: Project\\Intelligible Principal Component Analysis}

\usepackage[strings]{underscore}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\rhead{DS-GS 1013}
\lhead{Kho \& Peeters}
\chead{\thepage}

\setlength\parindent{0pt}

\begin{document}
\maketitle

We want to research different methods of retrieving intelligible principal components when looking at gene expressions. We consider the sparse principal component analysis model introduced by Zou, Hastie and Tibshirani \citeyear{Zou2006} the key piece of literature for this paper. After reading our paper, the reader should be informed about the structural differences between SPCA approaches, its potential applications, and benchmarks against other methods. As an application we will perform SPCA on UC Irvine's \textit{gene expression cancer RNA-seq data set} \cite{Dua2019} and interpret its results. We have the following hypotheses we want to check.
\begin{hyp}
SPCA is the primus inter pares of intelligible PCA methods.\footnote{We will be looking specifically at SPCA, SCoTLASS, thresholding of loadings and normal PCA.}
\end{hyp}
\begin{hyp}
SPCA is able to condense UC Irvine's \textit{gene expression cancer RNA-seq data set} to interpretable principal components revealing gene groupings.
\end{hyp}

\section*{Context}
Jolliffe \citeyear{Jolliffe1986} introduced the ubiquitous dimension-reduction technique PCA in 1986. It has since been a staple in statistics curricula and academic research. One of the most relevant drawbacks to using principal component analysis (henceforth PCA) is the difficulty of interpreting separate principal components as they are a linear combination of all the original variables. Since its creation, several attempts have been made at making the principal components more intelligible. One of those is the rotation of some or all the components to simplify them. However, this approach is only permissible on a certain subset of components, namely those with nearly equal variance \cite{Jolliffe1989, Jolliffe1995}.\\

Another approach frequently applied in order to increase the readability of the components is simply setting the loadings of variables with a small-magnitude to zero. This method could lead to misleading results for a series of reasons \cite{Cadima1995}. Vines \citeyear{Vines2000} devised a method that limits the individual loadings to $-1$, $0$ or $1$. Jolliffe, Trendafilov, and Uddin \citeyear{Jolliffe2003} introduced SCoTLASS, which is a modified PCA technique that is based on the least absolute shrinkage and selection operator (or lasso). Hence, penalizing small loadings on many different variables.\\

One of the later developments is sparse PCA (henceforth SPCA) developed by Zou, Hastie and Tibshirani \citeyear{Zou2006}. Similarly to SCoTLASS this method incorporates the lasso method, but in addition, they reformulate the PCA in a regression-type optimization problem. This method achieves overall sparse loadings with limited variable overlap between the different principal components. In the years following its first publication, SPCA has been improved upon in a plethora of ways \cite{Shen2008, Journee2010}.

\section*{Data set and methodology}
The UCI's dataset contains a random extraction of gene expressions of patients with different types of tumors, such as breast, kidney, cancer, colon, lung and prostate cancer. The data set has $801$ observations and $20531$ features which makes it interesting as an SPCA test case. We will use SPCA to group different gene expressions and their effect on the development of different types of cancer. Besides SPCA, we will also apply the SCoTLASS, thresholding of loadings and normal PCA methods and compare their principal components.

\pagebreak
\bibliographystyle{apacite}
\bibliography{references}
\end{document}

### 2020-04-05 Paper/Paper.tex
Updated paper and Main

@@
-\documentclass[12pt,letterpaper]{article}
+\documentclass[10pt,letterpaper]{article}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{graphicx}
+\usepackage{apacite}
+\usepackage{ntheorem}
+\newtheorem{hyp}{Hypothesis}
 \usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
+\author{
+	Kho, Lee\\
+	\textit{ltk224}
+	\and
+	Peeters, Jonas\\
+	\textit{jp5642}
+}
+\title{DS-GA 1013: Project\\Intelligible Principal Component Analysis}
+
+\usepackage[strings]{underscore}
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhf{}
+\rhead{DS-GS 1013}
+\lhead{Kho \& Peeters}
+\chead{\thepage}
+
+\setlength\parindent{0pt}
+
+\input{macros}
+
 \begin{document}
@@
 \end{enumerate}
+\section*{Simple Principal Components}
+\subsection*{Notes}
+Vines (2000) generalized this approach without restricting the integer set solution using a sequence of simplicity preserving transformations.
+Each transformation chooses a pair of vectors and rotates them orthogonally in such a way that the variance associated with the currently higher variance component of the pair is increased at the expense of the lower variance component
+\\\\
+This method what introduced by Vines \citeyear{Vines2000}.
+We transform the set of orthogonal axes.
+Essentially we will be orthogonally rotating and rescaling two orthogonal vectors $d_1$ and $d_2$.
+In general we can write an orthogonal rotation as $f_1 = Pd_1$ where
+\begin{equation}
+P=\begin{bmatrix}
+1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
+\end{bmatrix}
+\end{equation}
+The new variance-covariance matrix $V^*$ of the rotated data can be found as $V^*=P^TVP$ where $V$ is the variance-covariance matrix of the data before rotation. 
+\begin{align*}
+V=\begin{bmatrix}
+v_{11} & v_{12}\\v_{12} & v_{22}
+\end{bmatrix}\quad \text{and}\quad V^* &= P^TVP=\begin{bmatrix}
+1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
+\end{bmatrix}^T\begin{bmatrix}
+v_{11} & v_{12}\\v_{12} & v_{22}
+\end{bmatrix}\begin{bmatrix}
+1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
+\end{bmatrix}\\
+&= \begin{bmatrix}
+h
+\end{bmatrix}
+\end{align*}
+The variance in the direction $f_1$ can be found as follows:
+\begin{equation}
+\Var(Xf_1)=f_1^TVf_1=(Pd_1)^TVPd_1=d_1^TV^*d_1
+\end{equation}
+
+\pagebreak
+\bibliographystyle{apacite}
+\bibliography{references}
 \end{document}

### 2020-04-15 Paper/Archive.tex
Updated Paper for Simple Components

--- 2020-03-29 Paper/Paper.tex
@@
-\documentclass[12pt,letterpaper]{article}
-\usepackage[utf8]{inputenc}
-\usepackage[english]{babel}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{graphicx}
-\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
-\begin{document}
-\maketitle
-\section*{Deliverables}
-\begin{enumerate}
-\item Load the data
-\item Performance benchmarking
-\item PCA
-\item Thresholding PCA
-\item Simple Principal Components
-\item CPEV versus Sparsity plots
-\item SCoTLASS
-\item SPCA
-\end{enumerate}
-\end{document}
+\begin{equation}
+P=\begin{bmatrix}
+1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
+\end{bmatrix}
+\end{equation}
+The new variance-covariance matrix $V^*$ of the rotated data can be found as $V^*=P^TVP$ where $V$ is the variance-covariance matrix of the data before rotation. 
+\begin{align*}
+V=\begin{bmatrix}
+v_{11} & v_{12}\\v_{12} & v_{22}
+\end{bmatrix}\quad \text{and}\quad V^* &= P^TVP=\begin{bmatrix}
+1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
+\end{bmatrix}^T\begin{bmatrix}
+v_{11} & v_{12}\\v_{12} & v_{22}
+\end{bmatrix}\begin{bmatrix}
+1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
+\end{bmatrix}
+\end{align*}
+The variance in the direction $f_1$ can be found as follows:
+\begin{equation}
+\Var(Xf_1)=f_1^TVf_1=(Pd_1)^TVPd_1=d_1^TV^*d_1
+\end{equation}
+essentially we want to find a $\beta^*$ such that the variance $\upnu$ in the direction of the new direction $f_1$ is maximized.
+\begin{equation*}
+\upnu=\frac{\|d_1\|_2^2\upnu_{11}+2\beta\|d_1\|\|d_2\|\upnu_{12}+\beta^2\|d_2\|^2\upnu_{22}}{\|d_1\|^2+\beta^2\|d_2\|^2}
+\end{equation*}
+for which we find that $\upnu$ is maximal if:
+\begin{equation*}
+\beta^*=\frac{-\|d_1\|\|d_2\|(\upnu_{11}-\upnu_{22})+\sqrt{\|d_1\|^2\|d_2\|^2(\upnu_{11}-\upnu_{22})^2+4\|d_1\|^2\|d_2\|^2\upnu_{12}^2}}{2\|d_2\|^2\upnu_{12}}
+\end{equation*}

### 2020-04-15 Paper/Paper.tex
Updated Paper for Simple Components

--- 2020-03-29 Paper/Paper.tex
@@
 \usepackage{graphicx}
+\usepackage{upgreek}
+\usepackage{apacite}
+\usepackage{ntheorem}
+\newtheorem{hyp}{Hypothesis}
 \usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
+\author{
+	Kho, Lee\\
+	\textit{ltk224}
+	\and
+	Peeters, Jonas\\
+	\textit{jp5642}
+}
+\title{DS-GA 1013: Project\\Intelligible Principal Component Analysis}
+
+\usepackage[strings]{underscore}
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhf{}
+\rhead{DS-GS 1013}
+\lhead{Kho \& Peeters}
+\chead{\thepage}
+
+\setlength\parindent{0pt}
+
+\input{macros}
+
 \begin{document}
@@
 \end{enumerate}
+\section*{Simple Principal Components}
+Vines \citeyear{Vines2000} introduced the concept of simple principal component by restricting the magnitude of the loadings to integers such as $-1$, $0$ and $1$.
+This approached is based on the work for Hausman \citeyear{Hausman1982} who introduced a integer programming solution for assembling principal components using integers from a finite set.
+Each element is determined sequentially as to optimize the variance explained by each component.
+The problem with this method is that it rarely results in orthogonality over the components.\\
+\\
+Since it is hard to simplify principal components once they are calculated, Vines describes an alternative approach to obtaining simple components using an ortho-gonality preserving transformation on a pair of components.
+For each transformation of a pair of components, the variance of the data in the direction of one component is maximized while the variance in the other direction is minimized.
+This transformation is applied to a combination of principal components that has a high covariance.
+After this transformation, we exclude this pair of components for further transformations until all sensible transformations were made with the remaining components.\\
+\\
+During the "simplicity preserving transformation" introduced by Vines \citeyear{Vines2000}, we essentially orthogonally rotate and rescale two existing components $d_1$ and $d_2$.
+In general we can write an orthogonal rotation as $(f_1,f_2) = P(d_1,d_2)$
+\begin{equation}
+P=\begin{bmatrix}
+1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
+\end{bmatrix}
+\end{equation}
+V=\begin{bmatrix}
+v_{11} & v_{12}\\v_{12} & v_{22}
+\end{bmatrix}\quad \text{and}\quad V^* &= P^TVP
+
+\\
+\pagebreak
+\bibliographystyle{apacite}
+\bibliography{references}
 \end{document}

### 2020-04-15 Paper/Paper.tex
Added  Introduction, State of The Art and updated Simple Principal Components

@@
 \end{enumerate}
+
+\section*{Introduction}
+
+Principal component analysis (PCA) is a popular dimension-reduction technique that has wide applications across many fields.
+The goal of PCA is to reduce the dimension of the data while preserving as much variance as possible.
+PCA does this by computing orthogonal linear combinations of the original variables, called the principal components (PCs), such that the variance along these components are maximized.
+The resulting PCs are ordered from highest variance to lowest variance, and thus the first few principal components will retain the most variance in the data.\\
+\\
+One major limitation of PCA is that the resulting linear combinations of the original $p$ variables are difficult to interpret.
+This is due to the fact that, in most cases, the loadings of the resulting PCs are nonzero.
+This is particularly challenging when the number of variables $p$ is much larger than the number of examples $n$ (commonly called “High Dimension, Low Sample Size” data, or HDLSS), which is the case for many modern data sets.
+Consequently, much effort has gone into developing methods of approximating PCs with fewer than the $p$ original variables.\\
+\\
+In this paper, we will explore several sparse PCA methods that have been developed over the past few decades and discuss the benefits and drawbacks of each method.
+We will evaluate each method on a gene expression cancer RNA-Seq data set compiled by Samuele Fiorini at the University of Genoa \cite{Dua2019}.\\
+\\
+\section*{State of the Art}
+
+Initially, several simple methods of producing interpretable PCs were developed in the 1990s and early 2000s.
+Vines introduced a method of building simple PCs by restricting loadings to be from a small set of integers such as 1, 0 and -1.
+Another more \textit{ad hoc} way of producing sparse PCs is simple thresholding, as described by Cadima & Joliffe (1995).
+This approach sets loadings with absolute values below a certain threshold to zero.
+While both of these methods are relatively easy to implement and are often used in practice, the resulting PCs may be misleading.\\
+\\
+
 \section*{Simple Principal Components}
@@
 This transformation is applied to a combination of principal components that has a high covariance.
-After this transformation, we exclude this pair of components for further transformations until all sensible transformations were made with the remaining components.\\
+After this transformation, we exclude this pair of components for further transformations until all sensible transformations were made with the remaining components.
+We will describe the algorithm in the following paragraphs.\\
 \\
-During the "simplicity preserving transformation" introduced by Vines \citeyear{Vines2000}, we essentially orthogonally rotate and rescale two existing components $d_1$ and $d_2$.
-In general we can write an orthogonal rotation as $(f_1,f_2) = P(d_1,d_2)$
+For a data matrix $X\in\mathbb{R}^{N\times D}$e need to initialize $N$ principal components of length $D$ that are orthogonal to one another and consist out of integers.
+While does not propose initial components, we opt for a matrix $U$ constructed from [CONTINUE].
+First, we start by calculating the variances and covariances of the data in the directions of all of our components.
+Secondly, we transform the two principal components with the highest covariance update them.
+These components are excluded from consideration until all reasonable transformations with other components are completed.\\
+\\
+Two directions are transformed using the "simplicity preserving transformation" introduced by Vines \citeyear{Vines2000}.
+Essentially the transformation orthogonally rotates and rescales two existing components $d_1$ and $d_2$.
+In general we can write this orthogonal rotation as $(f_1,f_2) = P(d_1,d_2)$
 \begin{equation}
@@
 \end{equation}
-V=\begin{bmatrix}
-v_{11} & v_{12}\\v_{12} & v_{22}
-\end{bmatrix}\quad \text{and}\quad V^* &= P^TVP
+Similarly the covariance matrix $V^*$ of $f_1$ and $f_2$ can be written as the transformed covariance matrix $V$ of  $d_1$ and $d_2$:\footnote{Where $\Sigma$ is the covariance matrix of the data.}
+\begin{equation}
+V^* &= P^TVP\quad\text{with}\quad V = \begin{bmatrix}-& d_1 &-\\-&d_2&-\end{bmatrix}\Sigma\begin{bmatrix}\mid & \mid\\d_1^T & d_2^T\\\mid & \mid\end{bmatrix}
+\end{equation}
+When want this transformation $P(\beta)$ to yield a $V^*$ where the variance explained by the first direction $f_1$ is maximized at the cost of the variance explained by the second direction $f_2$ using the parameter $\beta$.
+Because we want to have integer vector before and after transformation we impose restrictions on $\beta^*$:
+\begin{equation}
+    \beta^*\in\bigg\{i/2^k, 2^k/i: i\in\{-2^k,-2^k+1,\ldots,2^k\}\text{ and }k\in\mathbb{Z}\bigg\}
+\end{equation}
+For example, if $k=1$, then $\beta^*\in\{-1, -1/2, 0, 1/2, 1\}$. Using this $\beta^*$ we retrieve the following new directions:
+\begin{equation}
+    \text{For }|\beta^*|\leq 1:\ \begin{cases}
+    f_1 = 2^kd_1 + 2^k\beta^*d_2\\
+    f_2 = 2^k\beta^*\|d_2\|^2-2^k\|d_1\|^2d_2
+    \end{cases}
+\end{equation}
+\begin{equation}
+    \text{For }|\beta^*|> 1:\ \begin{cases}
+    f_1 = 2^kd_1/\beta^* + 2^kd_2\\
+    f_2 = 2^k\|d_2\|^2-2^k\|d_1\|^2d_2/\beta^*
+    \end{cases}
+\end{equation}
 

### 2020-05-02 Paper/paper_archive0502.tex

\documentclass[12pt,letterpaper]{report}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{upgreek}
\usepackage{apacite}
\usepackage{ntheorem}
\usepackage{enumitem}
\usepackage{gensymb}
\graphicspath{ {./Figures/} }
\newtheorem{hyp}{Hypothesis}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\author{
	Kho, Lee\\
	\textit{ltk224}
	\and
	Peeters, Jonas\\
	\textit{jp5642}
}
\title{DS-GA 1013: Project\\Intelligible Principal Component Analysis}

\usepackage[strings]{underscore}
\usepackage{fancyhdr}
\fancypagestyle{plain}{
    \fancyhf{}
    \lhead{Kho \& Peeters}
    \chead{\thepage}
    \rhead{DS-GS 1013}
}
\pagestyle{plain}

\setlength\parindent{0pt}

\input{macros}

\begin{document}
\maketitle
% \section*{Outline}
% \begin{enumerate}
% \item Introduction\\
% \textit{Jonas: I feel like this part just needs some polishing, but everything we want it to do is already there.}
% \item State of the Art\\
% \textit{Overall this part should be kept for name-dropping and connecting the different papers based on their time line, core ideas and findings.}
% \begin{enumerate}
%     \item Models
%     \item Evaluation criteria
% \end{enumerate}
% \item Methodology
% \begin{enumerate}
%     \item Dataset 
%     \item Models
%     \begin{enumerate}
%         \item Thresholding
%         \item SCoT
%         \item SCoTLASS
%         \item Vines\\
%         \textit{Jonas: This one should be kept on the down low. We should spend minimal time explaining how it works and the results it yielded.}
%     \end{enumerate}
%     \item Evaluation criteria
%     \begin{enumerate}
%         \item Cumulative Explained Variance\\
%         \textit{Tables illustrating CEV versus components}
%         \item Statistical dispersion of explained variance of each principal components\\
%         \textit{Gini coefficient for each approach}
%         \item Metric for sparsity of our principal components\\
%         \textit{\cite{Hurley2008}, Check graph on conclusion's page. $\kappa_4$ seems to be a good contender since it heavily penalizes non-zero values. Maybe too aggressive, but we have a lot of choice in this paper.}
%         \item{Algorithm efficiency}
%     \end{enumerate}
% \end{enumerate}
% \item Results
% \begin{enumerate}
%     \item Table evaluation metrics per method of several hyper parameters
%     \item Interpretation of principal component for genes\\
%     \textit{Jonas: This might be challenging. I suggest we only do this for our "best" performing SPCA method.}
% \end{enumerate}
% \item Discussion
% \begin{enumerate}
%     \item Zhou's SPCA\\
%     \textit{If we don't have time to include it in our paper we can refer to it here and the findings in Zhou's paper.}
%     \item Shortly summarize the advantages and disadvantages or Sparse PCA in general and give a range of possible applications where Sparse PCA might be useful.
%     \item Enumerate interesting further research that might follow from this paper.
% \end{enumerate}
% \end{enumerate}

\section*{Introduction}

Principal component analysis (PCA) is a popular dimension-reduction technique that has wide applications across many fields. The goal of PCA is to reduce the dimension of the data while preserving as much variance as possible. PCA does this by computing orthogonal linear combinations of the original variables, called the principal components (PCs), such that the variance along these components are maximized.  The resulting PCs are ordered from highest variance to lowest variance, and thus the first few principal components will retain the most variance in the data.\\
\\
One major limitation of PCA is that the resulting linear combinations of the original $p$ variables are difficult to interpret. This is due to the fact that, in most cases, the loadings of the resulting PCs are nonzero. This is particularly challenging when the number of variables $p$ is much larger than the number of examples $n$ (commonly called “High Dimension, Low Sample Size” data, or HDLSS), which is the case for many modern data sets. Consequently, much effort has gone into developing methods of approximating PCs with fewer than the $p$ original variables.\\
\\
In this paper, we will explore several sparse PCA methods that have been developed over the past few decades and discuss the benefits and drawbacks of each method. We will evaluate each method on a breast tissue data set compiled by JP Marques de Sá at INEB-Instituto de Engenharia Biomédica.

\section*{State of the Art}
% PCA
Jolliffe \citeyear{Jolliffe1986} introduced the ubiquitous dimension-reduction technique PCA in 1986. It has since been a staple in statistics curricula and academic research. As previously mentioned, one of the most relevant drawbacks to using PCA is the difficulty of interpreting separate principal components as they are a linear combination of all the original variables. Since its creation, several attempts have been made at making the principal components more intelligible.\\
\\
% Thresholding
One of those approaches frequently applied is setting loadings with absolute values below a certain threshold to zero. This \textit{thresholding} method could lead to misleading results for a series of reasons as described by Cadima and Jolliffe \citeyear{Cadima1995}.
% Orthogonal rotations
Another common practice based on factor  analysis is to perform and orthogonal rotation on the regular PCA components \cite{Richman1986, Richman1987}. This rotation is then optimized by maximizing simplicity criteria such as \textit{varimax} and \textit{quartimax}. Both these criteria will move towards their maximum when the loadings of the vectors are either close to $0$ or $1$. The drawback of this method is that it does not preserve the successive maximization property inherent to PCA. Moreover Jolliffe \citeyear{Jolliffe1989, Jolliffe1995} warns that orthogonal rotations often lead to non-orthogonal loadings and pairwise correlation between the principal components, and should be reserved for principal components with near-equal variance.\\
\\
% SCoT
This rotation method is two-staged in nature as it requires you to first compute PCA and then rotate the resulting PCs optimally. However, Jolliffe and Uddin \citeyear{Jolliffe2000} introduced the one-stage \textit{SCoT} approach that combines the objectives of variance maximization (like PCA) and simplification (using the varimax criteria). Contrary to a the normal orthogonal rotation approach that stays within the subspace with maximum variance defined by the regular principal directions, SCoT's use of a penalty function will make the resulting principal components divert from this maximum variance subspace and hence faces a variance-simplicity trade-off.\\
\\
% Vines
While SCoT yields increased sparsity compared to PCA, the loadings will still be nonzero values. This is an issue that both Hausman \citeyear{Hausman1982} and Vines \citeyear{Vines2000} aim to solve. Hausman introduced an integer programming solution for assembling principal components using integers loadings from a finite set (typically {$-1$,$0$,$1$}). Each element is determined sequentially as to optimize the variance explained by each component.\\

%The problem with this method is that it rarely results in orthogonality over the components. Vines introduces a method based on simplicity and orthogonality preserving transformation on a pair of principal components. Essentially this transformation seeks to maximize the variance contained in one of the directions at cost of the other direction. From a starting set of orthogonal simple (integer) directions a pair of directions is selected based the highest absolute covariance. After the transformation this pair is restricted from further transformations until all trivial transformations are completed. The process can then start over eventually resulting in simple components where a few contain the most explained variance. Because of the nature of the transformation, the loadings of the resulting simple PCs will be restricted to a small set of integers such as $1$, $0$ and $-1$.\\
\\
% SCoTLASS
Jolliffe, Trendafilov, and Uddin \citeyear{Jolliffe2003} introduced \textit{SCoTLASS}, which is a modified PCA technique that is based on the least absolute shrinkage and selection operator (or LASSO). Including a LASSO penalty on the loadings helps to induce sparsity, however SCoTLASS is often computationally costly due to its non-convex nature. In addition, hyperparameter tuning for SCoTLASS can be challenging, altogether making SCoTLASS an impractical method for most applications.\\

% SPCA
One of the later developments is sparse PCA (henceforth SPCA) developed by Zou, Hastie and Tibshirani \citeyear{Zou2006}. Like SCoTLASS, this method incorporates the lasso method, but in addition, they reformulate the PCA in a regression-type optimization problem. This method achieves overall sparse loadings with limited variable overlap between the different principal components. However, unlike SCoT and SCoTLASS, SPCA results in non-orthogonal PCs. In the years following its first publication, SPCA has been improved upon in a plethora of ways \cite{Shen2008, Journee2010}.

%Initially, several simple methods of producing interpretable PCs were developed in the 1990s and early 2000s.  Vines  Another more \textit{ad hoc} way of producing sparse PCs is simple thresholding, as described by Cadima & Joliffe (1995).  While both of these methods are relatively easy to implement and are often used in practice, the resulting PCs may be misleading.\\
\\

% \section*{Simple Principal Components}
% Vines \citeyear{Vines2000} introduced the concept of simple principal component by restricting the magnitude of the loadings to integers such as $-1$, $0$ and $1$. This approached is based on the work for Hausman \citeyear{Hausman1982} who introduced a integer programming solution for assembling principal components using integers from a finite set. Each element is determined sequentially as to optimize the variance explained by each component. The problem with this method is that it rarely results in orthogonality over the components.\\
% \\
% Since it is hard to simplify principal components once they are calculated, Vines describes an alternative approach to obtaining simple components using an orthogonality preserving transformation on a pair of components. For each transformation of a pair of components, the variance of the data in the direction of one component is maximized while the variance in the other direction is minimized. This transformation is applied to a combination of principal components that has a high covariance. After this transformation, we exclude this pair of components for further transformations until all sensible transformations were made with the remaining components. We will describe the algorithm in the following paragraphs.\\
% \\
% For a data matrix $X\in\mathbb{R}^{N\times D}$e need to initialize $N$ principal components of length $D$ that are orthogonal to one another and consist out of integers. While does not propose initial components, we opt for a matrix $U$ constructed from [CONTINUE]. First, we start by calculating the variances and covariances of the data in the directions of all of our components. Secondly, we transform the two principal components with the highest covariance update them. These components are excluded from consideration until all reasonable transformations with other components are completed.\\
% \\
% Two directions are transformed using the "simplicity preserving transformation" introduced by Vines \citeyear{Vines2000}. Essentially the transformation orthogonally rotates and rescales two existing components $d_1$ and $d_2$. In general we can write this orthogonal rotation as $(f_1,f_2) = P(d_1,d_2)$ 
% \begin{equation}
% P=\begin{bmatrix}
% 1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
% \end{bmatrix}
% \end{equation}
% Similarly the covariance matrix $V^*$ of $f_1$ and $f_2$ can be written as the transformed covariance matrix $V$ of  $d_1$ and $d_2$:\footnote{Where $\Sigma$ is the covariance matrix of the data.}
% \begin{equation}
% V^* &= P^TVP\quad\text{with}\quad V = \begin{bmatrix}-& d_1 &-\\-&d_2&-\end{bmatrix}\Sigma\begin{bmatrix}\mid & \mid\\d_1^T & d_2^T\\\mid & \mid\end{bmatrix}
% \end{equation}
% When want this transformation $P(\beta)$ to yield a $V^*$ where the variance explained by the first direction $f_1$ is maximized at the cost of the variance explained by the second direction $f_2$ using the parameter $\beta$. Because we want to have integer vector before and after transformation we impose restrictions on $\beta^*$:
% \begin{equation}
%     \beta^*\in\bigg\{i/2^k, 2^k/i: i\in\{-2^k,-2^k+1,\ldots,2^k\}\text{ and }k\in\mathbb{Z}\bigg\}
% \end{equation}
% For example, if $k=1$, then $\beta^*\in\{-1, -1/2, 0, 1/2, 1\}$. Using this $\beta^*$ we retrieve the following new directions:
% \begin{equation}
%     \text{For }|\beta^*|\leq 1:\ \begin{cases}
%     f_1 = 2^kd_1 + 2^k\beta^*d_2\\
%     f_2 = 2^k\beta^*\|d_2\|^2-2^k\|d_1\|^2d_2
%     \end{cases}
% \end{equation}
% \begin{equation}
%     \text{For }|\beta^*|> 1:\ \begin{cases}
%     f_1 = 2^kd_1/\beta^* + 2^kd_2\\
%     f_2 = 2^k\|d_2\|^2-2^k\|d_1\|^2d_2/\beta^*
%     \end{cases}
% \end{equation}

\section*{Methodology}

%How did you tackle the question? How did you modify existing methods? What datasets did you use? What theoretical tools did you apply? If you deviated from your original plan, explain why.

\subsection*{Selected Sparse PCA Methods}

We implemented the following three methods to compare results across different sparse PCA approaches.\\
\begin{enumerate}[nolistsep]\\
    \item Simple Component Technique (SCoT)
    \item Simple Component Technique-LASSO (SCoTLASS)
    \item Sparse Principal Component Analysis (SPCA)\\
\end{enumerate}

Let $X \in \mathbb{R}^{n\times p}$ be a data set with $n$ samples and $p$ features and covariance matrix $\Sigma \in \mathbb{R}^{p\times p}$. SCoT seeks to solve the following constrained maximization problem for some regularization parameter $\lambda$ (here, $S(\mathbf{v})$ is the varimax rotation of a vector $\mathbf{v}$):

\begin{equation}
\begin{aligned}
\arg\max_{v_{i}} \quad & \mathbf{v_{i}^{T}}\Sigma \mathbf{v_{i}} + \lambda S(\mathbf{v_{i}})\\
\textrm{s.t.} \quad & \lVert \mathbf{v_{i}} \rVert_{2} = 1 \text{ and } \mathbf{v_{i}^{T}}\mathbf{v_{j}} = 0 \text{ for } j < i < p \\
\textrm{where} \quad & S(\mathbf{v_{i}}) = \frac{1}{p}\left[\sum_{k=1}^{p} \mathbf{v_{k}}^4- (\sum_{k=1}^{p} \mathbf{v_{k}}^2)^2 \right]\\ 
\end{aligned}
\end{equation}

The effect of including the varimax criterion in the SCoT objective function is that the solution seeks to maximize the variance of the squared loadings of the resulting components, driving loadings towards zero or $\pm{1}$. We optimized this SCoT objective using projected gradient ascent.\\

The SCoTLASS method solves the following constrained maximization problem for some regularization parameter $c$:

\begin{equation}
\begin{aligned}
\arg\max_{v_{i}} \quad & \mathbf{v_{i}^{T}}\Sigma \mathbf{v_{i}}\\
\textrm{s.t.} \quad & \lVert \mathbf{v_{i}} \rVert_{2} = 1 \text{, }
            \lVert \mathbf{v_{i}} \rVert_{1} < c \text{ and } \mathbf{v_{i}^{T}}\mathbf{v_{j}} = 0 \text{ for }  j < i < p \\
\end{aligned}
\end{equation}

Similarly to SCoT, we used projected gradient ascent to optimize SCoTLASS. However, the SCoTLASS problem involves an L1 penalty constraint, which is not differentiable at every point and thus poses an issue when using projected gradient ascent. \\

In order to remedy this, we approximated $\lVert \mathbf{v}\rVert_{1}$ as $\mathbf{v^T}\tanh{(\gamma \mathbf{v})}$ for a large $\gamma$ (specifically, $\gamma = 1000$) and reformatted the objective function to include a smooth exterior penalty function $P(x) = 0.5x(1+\tanh{(\gamma x)})$. This procedure is described in detail by Trendafilov \citeyear{Trendafilov2004}. These modifications result in the following adapted SCoTLASS optimization problem for some regularization parameter $t$ and large $\mu$ (we used $\mu = 1000$):

\begin{equation}
\begin{aligned}
\arg\max_{v_{i}} \quad & \mathbf{v_{i}^{T}}\Sigma \mathbf{v_{i}} - \mu P(\mathbf{v_{i}^T} \tanh{(\gamma\mathbf{v_{i}})} - t)\\
\textrm{s.t.} \quad & \lVert \mathbf{v_{i}} \rVert_{2} = 1 \text { and } \mathbf{v_{i}^{T}}\mathbf{v_{j}} = 0 \text{ for }  j < i < p \\
\end{aligned}
\end{equation}

SPCA solves the following constrained maximization problem.

\begin{equation}
\begin{aligned}
\arg\max_{v_{i}} \quad & \text{FILL IN}\\
\textrm{s.t.} \quad & \text{FILL IN}\\
\end{aligned}
\end{equation}

\subsection*{Data Set}

Although we originally planned on evaluating each sparse PCA method on a gene expression data set with over 20,000 variables, we decided to use a smaller data set to make it easier to interpret results. In addition, certain methods (such as SCoTLASS) are computationally intractable for high-dimension data sets such as gene expression data sets. Ultimately, we tested each method using a data set of measurements taken on recently excised breast tissue. The data set had 9 variables and 106 examples (see Table \ref{tab1:dta_vars} for descriptions of each variable). \\

\subsection*{Evaluation Criteria}

We evaluated these methods based on three criteria: sparsity, percentage of explained variance (PEV), and efficiency and scalability.\\

\textit{Sparsity:} Since the sparse PCA methods we evaluated often result in nonzero loadings that are very close to zero, we cannot simply count the number of zero loadings as a measure of sparsity. For this reason, we used the \textit{gini coefficient} to measure sparsity of loadings. The gini coefficient is a simple measure of weight distribution across a vector. The coefficient takes a value between 0 and 1 inclusive, with a 0 value indicating total equality of loadings and a value of 1 indicating total inequality (i.e. there is only one nonzero loading). Thus, a larger gini coefficient corresponds with higher sparsity. \\

\textit{Percentage of explained variance (PEV):} An effective sparse PCA method will achieve sparsity while maintaining as much explained variance in the first few principal components as possible. For each sparse PCA method, we calculated cumulative PEV (CPEV) of the PCs and plotted the results for comparison purposes. \\

\textit{Efficiency and scalability:} Certain sparse PCA methods work well with small data sets with few variables, but become computationally impractical with higher-dimensional data sets. For each method, we measured run time on the breast tissue data set and analyzed level of complexity to assess the suitability of each method for different data set types.\\

%Results: What results did you obtain? Do they make sense? Provide a thorough analysis. Negative results are completely fine (they can be very valuable!).

\section*{Results}

The table below gives a summary of the results of the different sparse PCA methods for varying regularization parameters.\\

[TABLE HERE]\\

We find that, as we begin to constrain loadings, both SCoT and SCoTLASS shift the majority of weight from variable 4 to variable 7. Variable 6 also surprisingly jumps from 6th largest weight in PCA to 3rd largest weight. This indicates that there may be more predictive significance in these two variables than regular PCs would suggest. \\

In terms of efficiency and runtime, the SPCA function takes considerably less time to run than SCoT and SCoTLASS. This is largely due to our use of projected gradient ascent in maximizing SCoT and SCoTLASS, which required very small step sizes over many iterations in order to for the algorithms to converge. SCoTLASS is particularly costly to optimize due to the objective function's nonconvexity. This results in possible local maxima, therefore requiring several random starting points to obtain the global maxima. \\

In optimizing SCoT and SCoTLASS, we found is quite challenging to select the best regularization parameter to yield optimal levels of sparsity and explained variance. There is little guidance from relevant literature on selecting regularization parameters for these methods.\\

In addition, as we varied the regularization parameter for SCoT, we also find that there is a sudden increase in loading sparsity, which is accompanied by a sharp decrease in explained variance. At $\lambda$ = 600, SCoT produces a PC 1 with one zero loading and most other loadings between 0.30 and 0.44 and explains approximately 60\% of variance in the data. At $\lambda$ = 650 however, the PC 1 that results has six zero loadings, with one of the three nonzero loadings carrying nearly all the weight with a 0.98 coefficient, and explains only 20\% of variance in the data. This phenomenon makes SCoT an even more challenging model to tune.

% Pros of SCoT:
% - PCs are orthogonal
% - Computationally simpler than SCoTLASS
% Cons of SCoT:
% - difficult to tune 
%
% Pros of SCoTLASS:
% - PCs are orthogonal
% Cons of SCoTLASS:
% - Costly, requires random starts
% - difficult to tune
% - resulting PCs not as sparse as those resulting from other sparse PCA methods
% - selects at most n zero loadings (problematic when p >> n)
% Pros of SPCA:
%
% Cons of SPCA:
% - produces non orthogonal pcs

\section*{Discussion}

%Discussion: What did you find out? Do the results suggest any other interesting questions to explore?

% - explore ways to create uncorrelated sparse PCAs
% - explore other optimization methods for SCoT and SCoTLASS that are more efficient

% References
\pagebreak
\bibliographystyle{apacite}
\bibliography{references}

% Appendix
\pagebreak
\appendix
\section*{Appendix}
\begin{table}[htpb!]
\caption{Variable descriptions}
\label{tab1:dta_vars}
\begin{center}
 \begin{tabular}{||c l||}
 \hline
Variable & Variable Description \\ [0.5ex] 
 \hline\hline
 $x_{1}$ & I0 - impedivity (ohm) at zero frequency \\ 
 \hline
 $x_{2}$ & 	PA500 - phase angle at 500 KHz \\
 \hline
 $x_{3}$ & HFS - high-frequency slope of phase angle \\
 \hline
 $x_{4}$ & DA - impedance distance between spectral ends \\
 \hline
 $x_{5}$ & AREA - area under spectrum \\
 \hline
 $x_{6}$ & 	A/DA - area normalized by DA \\
 \hline
 $x_{7}$ & MAX IP - maximum of the spectrum \\
 \hline
 $x_{8}$ & 	DR - distance between I0 and real part of the maximum frequency point \\
 \hline
 $x_{9}$ & P - length of the spectral curve \\ [1ex] 
 \hline
\end{tabular}
\end{center}
\end{table}
\end{document}

### 2020-05-02 Presentation/Presentation.tex
Added beamer

\documentclass[11pt]{beamer}
\usetheme{Warsaw}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\author{
	Kho, Lee
	\and
	Peeters, Jonas
}
\title{Sparse Principal Components Analysis}
%\setbeamercovered{transparent} 
%\setbeamertemplate{navigation symbols}{} 
%\logo{} 
\institute{\includegraphics{../Figures/nyu_long_black.png}} 
\date{\today} 
\subject{DS-GA 1013}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}\frametitle{Introduction}
\begin{itemize}
\item 
\end{itemize}
\end{frame}

\begin{frame}\frametitle{State of the Art}
\begin{itemize}
\item 
\end{itemize}
\end{frame}

\begin{frame}\frametitle{Methodology}
\begin{itemize}
\item 
\end{itemize}
\end{frame}

\begin{frame}\frametitle{Results}
\begin{itemize}
\item 
\end{itemize}
\end{frame}

\begin{frame}\frametitle{Discussion}
\begin{itemize}
\item 
\end{itemize}
\end{frame}

\end{document}

### 2020-05-02 Paper/lit_review_JP.tex
Updates from Overleaf

\documentclass{article}
\begin{document}
\section{Jonas literature study 04/22}
\begin{enumerate}
    \item \cite{Berthet2013}
    \begin{enumerate}
         \item They devise a minimax optimal test that relies on the so-called k-sparse largest eigenvalue of the empirical covariance matrix. It captures the largest amount of empirical variance explained by any k of the original variables. The test tries to detect sparse principal components.
        \item Classical PCA produces inconsistent estimators of the directions that explain the most variance. \cite{Johnstone2009} (and more in paper)
        \item Johnstone \citeyear{Johnstone2001} introduces the spiked covariance model which provides a natural setting for statistical problems. The model relies on the assumption that there exists a small number of directions that explain most of the variance. Data is drawn from a multi-Gaussian distribution with mean zero and covariance matrix given by $I + \theta vv^T$, v is a unit norm sparse vector.
    \end{enumerate}
    \item \cite{Johnstone2001}
    \begin{enumerate}
        \item Too complex don't discuss
    \end{enumerate}
    \item \cite{Johnstone2009}
    \begin{enumerate}
        \item Some initial reduction in dimensionality is desirable before applying PCA.
        \item The initial reduction in dimensionality is best achieved by working in a basis in which the signals have a sparse representation.
        \item Overall too dense to read properly
    \end{enumerate}
    \item \cite{Wang2016}
    \begin{enumerate}
        \item Sparse PCA is employed to identify a small number foe interpretable directions that represent the data succinctly.
        \item Regular PCA works well when $n >> p$, but starts faltering when $p > n$ (see also \cite{Johnstone2009}).
        \item Since the meaningfulness of the components depends on the characteristics of the problem (sample size, dimensionality, sparsity level and signal-to-noise ratio).
        \item \cite{Ma2013} shows that his estimator for the SPCA attains the minimax rate of convergence over a certain Gaussian class of distributions, provided that the sparsity parameter $k$ is treated as a fixed constant. \cite{Cai2013, Vu2013} allows for $k$ to vary depending on the sample size $n$ of the problem. These papers settle the question of sparse principal component estimation. However they are not computable in polynomial time.
        \item \cite{Berthet2013} made improvement in this computability problem by testing versus an alternative hypothesis.
        \item They themselves also try to improve computability.
    \end{enumerate}
    \item \cite{Vu2013}
    \begin{enumerate}
        \item Study SPCA in high dimensions where $p>>n$ and introduce to complementary notions: row sparsity and column sparsity.
        \item Problem: whether SPCA methods can optimally estimate the subspace spanned by the leading eigenvectors, that is, the principal subspace of variation.
        \item \cite{Ma2013} proposed a sparse principal subspace estimator based on iterative thresholding, and derived its rate of convergence under a spiked covariance model.
        similar to \cite{Birnbaum2013}.
        \item The use of the spiked covariance model is warranted by two reasons. First of all, it simplifies analyses and enables the exploitation of special properties of the multivariate Gaussian distribution. The second is that it excludes the possibility of the variables having equal variances.
    \end{enumerate}
    \item \cite{Witten2009}
    \begin{enumerate}
        \item Present a penalized matrix decomposition (PMD) framework for computing a rank-$K$ approximation of a matrix. The matrix $X$ is approximated as $\hat{X}=\sum_{k=1}^K d_k u_k v_k^T$ where $d_k$, $u_k$ and $v_k$ minimize the squared Frobenius norm of $X-\hat{X}$ subject to penalities on $u_k$ and $v_k$. They find that if they apply $\ell_1$-penalities on $v_k$ it yields an efficient algorithm for the SCoTLASS proposal \cite{Jolliffe2003}.
    \end{enumerate}
\end{enumerate}

\bibliographystyle{apacite}
\bibliography{references}
\end{document}

### 2020-05-03 Scripts/sparse_pca.py
Started the Results notebook

@@ -1,0 +1,633 @@
+import pandas as pd
+import numpy as np
+from sklearn.decomposition import PCA
+from sklearn import preprocessing
+from sklearn.preprocessing import StandardScaler
+import matplotlib.pyplot as plt
+from scipy.optimize import minimize
+from scipy.linalg import null_space
+from IPython.display import clear_output
+from sklearn.linear_model import LassoLars
+import random
+import warnings
+warnings.filterwarnings('ignore')
+
+def gini(array):
+    """Calculate the Gini coefficient of a numpy array."""
+    # based on bottom eq:
+    # http://www.statsdirect.com/help/generatedimages/equations/equation154.svg
+    # from:
+    # http://www.statsdirect.com/help/default.htm#nonparametric_methods/gini.htm
+    # All values are treated equally, arrays must be 1d:
+    array = array.flatten()
+    if np.amin(array) < 0:
+        # Values cannot be negative:
+        array -= np.amin(array)
+    # Values cannot be 0:
+    array += 0.0000001
+    # Values must be sorted:
+    array = np.sort(array)
+    # Index per array element:
+    index = np.arange(1,array.shape[0]+1)
+    # Number of array elements:
+    n = array.shape[0]
+    # Gini coefficient:
+    return ((np.sum((2 * index - n  - 1) * array)) / (n * np.sum(array)))
+
+def reg_PCA(X, k = "all"): 
+    """
+    function takes an n x p feature matrix
+    returns two arrays:
+    - array with percentage of explained variance in first k principal directions (k_comp x 1)
+    - array with principal directions (k_comp x p)
+    """
+    X = StandardScaler().fit_transform(X)
+    if k == "all": k = min(X.shape[0],X.shape[1])
+    pca = PCA(n_components = k)
+    pca.fit(X)
+    PEVs = pca.explained_variance_ratio_
+    prin_comp = pca.components_
+    EVs = pca.explained_variance_
+    
+    return PEVs, prin_comp, EVs
+
+def threshold_PCA(X, thresh = 1e-1, k = "all"):
+    """
+    function takes
+    - X: n x p feature matrix
+    - thresh: float representing this non-zero cutoff
+    - k: integer for number of principal directions wanted
+    returns one array:
+    - array with principal components in its columns (k x p)
+    """
[…]

### 2020-05-04 Paper/Paper.tex
Updates from Overleaf

--- 2020-04-05 Paper/Paper.tex
@@
-\documentclass[10pt,letterpaper]{article}
+\documentclass[11pt,letterpaper]{report}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{graphicx}
+\usepackage{xcolor}
+\usepackage{upgreek}
 \usepackage{apacite}
 \usepackage{ntheorem}
+\usepackage{enumitem}
+\usepackage{gensymb}
+\usepackage[ruled,vlined, linesnumbered]{algorithm2e}
+\graphicspath{ {./Figures/} }
 \newtheorem{hyp}{Hypothesis}
@@
 \usepackage{fancyhdr}
-\pagestyle{fancy}
+\fancypagestyle{plain}{
     \fancyhf{}
-\rhead{DS-GS 1013}
     \lhead{Kho \& Peeters}
     \chead{\thepage}
+    \rhead{DS-GS 1013}
+}
+\pagestyle{plain}
 
 \setlength\parindent{0pt}
+\allowdisplaybreaks
 
@@
 \maketitle
-\section*{Deliverables}
-\begin{enumerate}
-\item Load the data
-\item Performance benchmarking
-\item PCA
-\item Thresholding PCA
-\item Simple Principal Components
-\item CPEV versus Sparsity plots
-\item SCoTLASS
-\item SPCA
-\end{enumerate}
-\section*{Simple Principal Components}
-\subsection*{Notes}
-Vines (2000) generalized this approach without restricting the integer set solution using a sequence of simplicity preserving transformations.
-Each transformation chooses a pair of vectors and rotates them orthogonally in such a way that the variance associated with the currently higher variance component of the pair is increased at the expense of the lower variance component
-\\\\
-This method what introduced by Vines \citeyear{Vines2000}.
-We transform the set of orthogonal axes.
-Essentially we will be orthogonally rotating and rescaling two orthogonal vectors $d_1$ and $d_2$.
-In general we can write an orthogonal rotation as $f_1 = Pd_1$ where
+\section*{Introduction}
+
+Principal component analysis (PCA) is a popular dimension-reduction technique that has wide applications across many fields.
+The goal of PCA is to reduce the dimension of the data while preserving as much variance as possible.
+PCA does this by computing orthogonal linear combinations of the original variables, called the principal components (PCs), such that the variance along these components are maximized.
+The resulting PCs are ordered from highest variance to lowest variance, and thus the first few principal components will retain the most variance in the data.\\
+\\
+One major limitation of PCA is that the resulting linear combinations of the original $p$ variables are difficult to interpret.
+This is due to the fact that, in most cases, the loadings of the resulting PCs are nonzero.
+This is particularly challenging when the number of variables $p$ is much larger than the number of examples $n$ (commonly called “High Dimension, Low Sample Size” data, or HDLSS), which is the case for many modern data sets.
+Consequently, much effort has gone into developing methods of approximating PCs with fewer than the $p$ original variables.\\
+\\
+In this paper, we will explore several sparse PCA methods that have been developed over the past few decades and discuss the benefits and drawbacks of each method.
+We will evaluate each method on a breast tissue data set compiled by JP Marques de Sá at INEB-Instituto de Engenharia Biomédica.
+
+\section*{State of the Art}
+% PCA
+Jolliffe \citeyear{Jolliffe1986} introduced the ubiquitous dimension-reduction technique PCA in 1986.
+It has since been a staple in statistics curricula and academic research.
+As previously mentioned, one of the most relevant drawbacks to using PCA is the difficulty of interpreting separate principal components as they are a linear combination of all the original variables.
+Since its creation, several attempts have been made at making the principal components more intelligible.\\
+\\
+% Thresholding
+One of those approaches frequently applied is setting loadings with absolute values below a certain threshold to zero.
+This \textit{thresholding} method could lead to misleading results for a series of reasons as described by Cadima and Jolliffe \citeyear{Cadima1995}.
+% Orthogonal rotations
+Another common practice based on factor  analysis is to perform and orthogonal rotation on the regular PCA components \cite{Richman1986, Richman1987}.
+This rotation is then optimized by maximizing simplicity criteria such as \textit{varimax} and \textit{quartimax}.
+Both these criteria will move towards their maximum when the loadings of the vectors are either close to $0$ or $1$.
+The drawback of this method is that it does not preserve the successive maximization property inherent to PCA.
+Moreover Jolliffe \citeyear{Jolliffe1989, Jolliffe1995} warns that orthogonal rotations often lead to non-orthogonal loadings and pairwise correlation between the principal components, and should be reserved for principal components with near-equal variance.\\
+\\
+% SCoT
+This rotation method is two-staged in nature as it requires you to first compute PCA and then rotate the resulting PCs optimally.
+However, Jolliffe and Uddin \citeyear{Jolliffe2000} introduced the one-stage \textit{SCoT} approach that combines the objectives of variance maximization (like PCA) and simplification (using the varimax criteria).
+Contrary to a the normal orthogonal rotation approach that stays within the subspace with maximum variance defined by the regular principal directions, SCoT's use of a penalty function will make the resulting principal components divert from this maximum variance subspace and hence faces a variance-simplicity trade-off.\\
+\\
+% Vines
+While SCoT yields increased sparsity compared to PCA, the loadings will still be nonzero values.
+This is an issue that both Hausman \citeyear{Hausman1982} and Vines \citeyear{Vines2000} aim to solve.
+Hausman introduced an integer programming solution for assembling principal components using integers loadings from a finite set (typically {$-1$,$0$,$1$}).
+Each element is determined sequentially as to optimize the variance explained by each component.\\
+\\
+% SCoTLASS
+Jolliffe, Trendafilov, and Uddin \citeyear{Jolliffe2003} introduced \textit{SCoTLASS}, which is a modified PCA technique that is based on the least absolute shrinkage and selection operator (or LASSO).
+Including a LASSO penalty on the loadings helps to induce sparsity, however SCoTLASS is often computationally costly due to its non-convex nature.
+In addition, hyperparameter tuning for SCoTLASS can be challenging, altogether making SCoTLASS an impractical method for most applications.\\
+
+% SPCA
+One of the later developments is sparse PCA (henceforth SPCA) developed by Zou, Hastie and Tibshirani \citeyear{Zou2006}.
+Like SCoTLASS, this method incorporates the lasso method, but in addition, they reformulate the PCA in a regression-type optimization problem.
+This method achieves overall sparse loadings with limited variable overlap between the different principal components.
+However, unlike SCoT and SCoTLASS, SPCA results in non-orthogonal PCs.
+In the years following its first publication, SPCA has been improved upon in a plethora of ways \cite{Shen2008, Journee2010}.\\
+\\
+% After SPCA
+Developments in the field of sparse PCA after the introduction of SPCA mainly build on the spiked covariance model introduced by Johnstone \citeyear{Johnstone2001}.
+The model assumes that there exist a small number of natural directions that explains most of the variance.
+It states that the data is drawn from a multi-Gaussian distribution with zero-mean and a covariance matrix given by $I+\theta vv^T$, where $v$ is a unit norm sparse vector.
+A significant chunk of the literature is devoted to finding this $v$ in high-dimensionality settings where the number of features $p$ is way larger than the number of observations $n$.
+In 2013, \citeauthor{Ma2013} presents an estimator for Sparse PCA that attains the convergence, but is confronted with computational bottleneck as the algorithm is not computable in polynomial time.
+Further research focuses on solving this computational problem\footnote{Such as \citeA{Wang2016, Cai2013, Vu2013, Berthet2013}.}.
+
+\section*{Methodology}
+
+%How did you tackle the question? How did you modify existing methods? What datasets did you use? What theoretical tools did you apply? If you deviated from your original plan, explain why.
+
+\subsection*{Simple Component Technique (SCoT)}
+The Simple Component Technique introduced by Jolliffe and Uddin \cite{Jolliffe2000} optimizes a combination of both large variance and simplicity in one stage.
+Every component is also constrained to be orthogonal and uncorrelated with every other component.
+The SCoT maximization problem can be written as:
+\begin{equation*}
+\begin{aligned}
+\arg\max_{v_{i}} \quad & \underbrace{v_{i}^{T}\Sigma v_{i}}_{\text{Variance}} + \lambda \underbrace{S(v_{i})}_{\text{Sparsity}}\\
+\textrm{s.t.} \quad & \lVert v_{i} \rVert_{2} = 1 \text{ and } v_{i}^{T}v_{j} = 0 \text{ for } j < i < p \\
+\textrm{where} \quad & S(v_{i}) = \frac{1}{p}\left[\sum_{k=1}^{p} v_{k}^4- (\sum_{k=1}^{p} v_{k}^2)^2 \right]\\ 
+\end{aligned}
+\end{equation*}
+Where $\Sigma\in\mathbb{R}^{n \times p}$ is the covariance matrix of the zero-mean dataset $X \in \mathbb{R}^{n\times p}$ with $n$ samples and $p$ features.
+$S(v_i)$ is often referred to as the varimax rotation criteria and simplifies the structure of vector $v_i$ through driving the several loading to zero by maximizing the variance of the squared vector loadings.
+The important of this sparsity components in the maximization problem is regulated by the term $\lambda$.
+Since both the variance and sparsity terms in the maximization problem are easily differentiable, we chose to implement this method through the use of gradient descent.
+
+\subsection*{Simple Component Technique-LASSO (SCoTLASS)}
+The SCoTLASS method solves the following constrained maximization problem for some regularization parameter $c$:
+
 \begin{equation}
-P=\begin{bmatrix}
-1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
-\end{bmatrix}
-\end{equation}
-The new variance-covariance matrix $V^*$ of the rotated data can be found as $V^*=P^TVP$ where $V$ is the variance-covariance matrix of the data before rotation. 
-\begin{align*}
-V=\begin{bmatrix}
-v_{11} & v_{12}\\v_{12} & v_{22}
-\end{bmatrix}\quad \text{and}\quad V^* &= P^TVP=\begin{bmatrix}
-1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
-\end{bmatrix}^T\begin{bmatrix}
-v_{11} & v_{12}\\v_{12} & v_{22}
-\end{bmatrix}\begin{bmatrix}
-1 & \|d_2\|_2^2 \beta\\\beta & -\|d_1\|_2^2
-\end{bmatrix}\\
-&= \begin{bmatrix}
-h
-\end{bmatrix}
-\end{align*}
-The variance in the direction $f_1$ can be found as follows:
-\begin{equation}
-\Var(Xf_1)=f_1^TVf_1=(Pd_1)^TVPd_1=d_1^TV^*d_1
+\begin{aligned}
+\arg\max_{v_{i}} \quad & v_{i}^{T}\Sigma v_{i}\\
+\textrm{s.t.} \quad & \lVert v_{i} \rVert_{2} = 1 \text{, }
+            \lVert v_{i} \rVert_{1} < c \text{ and } v_{i}^{T} v_{j} = 0 \text{ for }  j < i < p \\
+\end{aligned}
 \end{equation}
 
+Similarly to SCoT, we used projected gradient ascent to optimize SCoTLASS.
+However, the SCoTLASS problem involves an L1 penalty constraint, which is not differentiable at every point and thus poses an issue when using projected gradient ascent.\\
+\\
+In order to remedy this, we approximated $\lVert v\rVert_{1}$ as $v^T\tanh{(\gamma v)}$ for a large $\gamma$ (specifically, $\gamma = 1000$) and reformatted the objective function to include a smooth exterior penalty function $P(x) = 0.5x(1+\tanh{(\gamma x)})$.
+This procedure is described in detail by Trendafilov \citeyear{Trendafilov2004}.
+These modifications result in the following adapted SCoTLASS optimization problem for some regularization parameter $t$ and large $\mu$ (we used $\mu = 1000$):
+
+\begin{equation}
+\begin{aligned}
+\arg\max_{v_{i}} \quad & v_{i}^{T}\Sigma v_{i} - \mu P(v_{i}^T \tanh{(\gamma v_{i})} - t)\\
+\textrm{s.t.} \quad & \lVert v_{i} \rVert_{2} = 1 \text { and } v_{i}^{T}v_{j} = 0 \text{ for }  j < i < p \\
+\end{aligned}
+\end{equation}
+
+\subsection*{Sparse Principal Components (SPCA)}
+SPCA was introduced by Zou et al.
+\cite{Zou2006} and applies the elastic net to find principal components with sparse loadings.
+They start form the regular PCA principal components of the zero-mean dataset $X$ with as SVD decomposition $UDV^T$ where $Z=UD$ are the uncorrelated and orthogonal principal components.
+SPCA will attempt to find a $v_i$ that is a sparse estimate of the $i$th principal component $Z_i = U_iD_{ii}$ regularized with an elastic net.
+This estimate $v_i$ can be found solving the following maximization problem:
+\begin{equation*}
+\arg\max_{v_{i}} \quad \|Z_i - X v_i\|^2 + \lambda\|v_i\|^2 + \lambda_1\|v_i\|_1
+\end{equation*}
+If we look at this equation we recognize a connection between PCA and the regression method.
+The addition of the positive ridge penalty $\lambda\|v_i\|^2$ ensures that even in the case of a $X$ where is not full-rank because $p>>n$ we will find a unique solution for $v_i$.
+Zou et al.
+continue with this similarity between PCA and a elastic net regression and derive the following optimization problem for the $k$ first principal components:
+\begin{equation*}
+\begin{aligned}
+\arg\max_{B} \quad & \sum_{i=1}^{k} \left(\|X\alpha_j-X\beta_j\|^2 + \lambda\|\beta_j\|^2 +\lambda_{1,j} \|\beta_j\|_1\right)\\
+\textrm{s.t.} \quad & \|\alpha\|^2 = 1\\
+\end{aligned}
+\end{equation*}
+with $A_{p\times k} = [\alpha_1,\ldots,\alpha_k]$ where $\alpha_i$ initially equals the $i$th principal component $Z_i$ and $B_{p\times k} = [\beta_1,\ldots,\beta_k]$ where $\beta_j\propto Z_i$.
+In order to numerically solve this maximization problem we first look for $B$ given $A$ and then $A$ given $B$ in an alternating fashion until $A$ converges.
+The general SPCA algorithm looks like:\vspace*{1mm}\\
+\begin{algorithm}[H]
+\SetAlgoLined
+\KwResult{$V = [v_1,\ldots,v_k]$ where $\|v_j\|=1$}
+$A_0 = Z$\;
+\While{$\|A_{t+1} - A_{t}\|_F \ne 0$}{
+    \For{$\beta_j$ in $B_{t+1}$}{
+    $Y_j = X\alpha_j$\;
+    $\beta_j = \arg\min_\beta \|Y_j - X\beta_j\|^2 + \lambda\|\beta\|^2 + \lambda_{1,j}\|\beta\|_1$\;
+    }
+    $X^TXB_{t+1}=UDV^T$\;
+    $A_{t+1} = UV^T$\;
+}
+\For{$\alpha_j$ in $A$}{
+$v_j = \alpha_j/\|\alpha_j\|$\;
+}
+\caption{General SPCA Algorithm}
+\label{algo1:SPCA}
+\end{algorithm}\vspace*{1mm}
+In order to solve line $5$ in the algorithm we apply the LARS-EN method for elastic net regressions introduced by Zou and Hastie \cite{Zou2005} which is itself an adaptation of Efron's \citeyear{Efron2004} LARS method.
+Essentially, we are able to transform an elastic net problem to a simple lasso problem and apply LARS by constructing an artificial data matrices $X^*$ and $Y^*$ based on $X$, $Y$ and $\lambda$.
+
+\subsection*{Data Set}
+
+Although we originally planned on evaluating each sparse PCA method on a gene expression data set with over 20,000 variables, we decided to use a smaller data set to make it easier to interpret results.
+In addition, certain methods (such as SCoTLASS) are computationally intractable for high-dimension data sets such as gene expression data sets.
+Ultimately, we tested each method using a data set of measurements taken on recently excised breast tissue.
+The data set had 9 variables and 106 examples (see Table \ref{tab1:dta_vars} for descriptions of each variable).
+\\
+
+\subsection*{Evaluation Criteria}
+
+We evaluated these methods based on three criteria: sparsity, percentage of explained variance (PEV), and efficiency and scalability.\\
+
+\textit{Sparsity:} Since the sparse PCA methods we evaluated often result in nonzero loadings that are very close to zero, we cannot simply count the number of zero loadings as a measure of sparsity.
+For this reason, we used the \textit{gini coefficient} to measure sparsity of loadings as recommended by \citeA{Hurley2008}.
+The gini coefficient is a simple measure of weight distribution across a vector.
+The coefficient takes a value between 0 and 1 inclusive, with a 0 value indicating total equality of loadings and a value of 1 indicating total inequality (i.e.
+there is only one nonzero loading).
+Thus, a larger gini coefficient corresponds with higher sparsity.
+\\
+
+\textit{Percentage of explained variance (PEV):} An effective sparse PCA method will achieve sparsity while maintaining as much explained variance in the first few principal components as possible.
+For each sparse PCA method, we calculated cumulative PEV (CPEV) of the PCs and plotted the results for comparison purposes.
+\\
+
+\textit{Efficiency and scalability:} Certain sparse PCA methods work well with small data sets with few variables, but become computationally impractical with higher-dimensional data sets.
+For each method, we measured run time on the breast tissue data set and analyzed level of complexity to assess the suitability of each method for different data set types.\\
+
+%Results: What results did you obtain? Do they make sense? Provide a thorough analysis. Negative results are completely fine (they can be very valuable!).
+
+\section*{Results}
+
+The table below gives a summary of the results of the different sparse PCA methods for varying regularization parameters.\\
+
+{\color{red}[TABLE HERE]
+We want four rows: PCA, SCoT, SCoTLASS, SPCA for their "best" specification.
+We will show the difference within each model in tables and figures in the appendix.
+Columns containing: Weighted Gini coefficient, distribution of explained variance, }
+\\
+
+We find that, as we begin to constrain loadings, both SCoT and SCoTLASS shift the majority of weight from variable 4 to variable 7.
+Variable 6 also surprisingly jumps from 6th largest weight in PCA to 3rd largest weight.
+This indicates that there may be more predictive significance in these two variables than regular PCs would suggest.
+\\
+
+In terms of efficiency and runtime, the SPCA function takes considerably less time to run than SCoT and SCoTLASS.
+This is largely due to our use of projected gradient ascent in maximizing SCoT and SCoTLASS, which required very small step sizes over many iterations in order to for the algorithms to converge.
+SCoTLASS is particularly costly to optimize due to the objective function's nonconvexity.
+This results in possible local maxima, therefore requiring several random starting points to obtain the global maxima.
+\\
+
+In optimizing SCoT and SCoTLASS, we found is quite challenging to select the best regularization parameter to yield optimal levels of sparsity and explained variance.
+There is little guidance from relevant literature on selecting regularization parameters for these methods.\\
+
+In addition, as we varied the regularization parameter for SCoT, we also find that there is a sudden increase in loading sparsity, which is accompanied by a sharp decrease in explained variance.
+At $\lambda$ = 600, SCoT produces a PC 1 with one zero loading and most other loadings between 0.30 and 0.44 and explains approximately 60\% of variance in the data.
+At $\lambda$ = 650 however, the PC 1 that results has six zero loadings, with one of the three nonzero loadings carrying nearly all the weight with a 0.98 coefficient, and explains only 20\% of variance in the data.
+This phenomenon makes SCoT an even more challenging model to tune.
+
+\section*{Discussion}
+%Discussion: What did you find out? Do the results suggest any other interesting questions to explore?
+
+% Pros of SCoT:
+% - PCs are orthogonal
+% - Computationally simpler than SCoTLASS
+% Cons of SCoT:
+% - difficult to tune 
+%
+% Pros of SCoTLASS:
+% - PCs are orthogonal
+% Cons of SCoTLASS:
+% - Costly, requires random starts
+% - difficult to tune
+% - resulting PCs not as sparse as those resulting from other sparse PCA methods
+% - selects at most n zero loadings (problematic when p >> n)
+% Pros of SPCA:
+%
+% Cons of SPCA:
+% - produces non orthogonal pcs
+
+% Discussion points:
+% - explore ways to create uncorrelated sparse PCAs
+% - explore other optimization methods for SCoT and SCoTLASS that are more efficient
+% - test methods with higher dimensional data sets
+
+% References
 \pagebreak
@@
 \bibliography{references}
+
+% Appendix
+\pagebreak
+\appendix
+\section*{Appendix}
+\begin{table}[htpb!]
+\caption{Variable descriptions}
+\label{tab1:dta_vars}
+\begin{center}
+ \begin{tabular}{||c l||}
+ \hline
+Variable & Variable Description \\ [0.5ex] 
+ \hline\hline
+ $x_{1}$ & I0 - impedivity (ohm) at zero frequency \\ 
+ \hline
+ $x_{2}$ & 	PA500 - phase angle at 500 KHz \\
+ \hline
+ $x_{3}$ & HFS - high-frequency slope of phase angle \\
+ \hline
+ $x_{4}$ & DA - impedance distance between spectral ends \\
+ \hline
+ $x_{5}$ & AREA - area under spectrum \\
+ \hline
+ $x_{6}$ & 	A/DA - area normalized by DA \\
+ \hline
+ $x_{7}$ & MAX IP - maximum of the spectrum \\
+ \hline
+ $x_{8}$ & 	DR - distance between I0 and real part of the maximum frequency point \\
+ \hline
+ $x_{9}$ & P - length of the spectral curve \\ [1ex] 
+ \hline
+\end{tabular}
+\end{center}
+\end{table}
+
+\begin{table}[htpb!]
+    \centering
+    \caption{SCoT Results for different parameters}
+    \label{rstl:scot}
+    \include{../Tables/PCA.tex}
+\end{table}
+
 \end{document}

### 2020-05-04 Scripts/sparse_pca.py
Added tables per method

@@ -342,7 +342,7 @@ def scotlassGradAsc(sigma, V, reg_param , x0 = 'default',
     return v
 
 def SCoTLASS(X, reg_param, x0 = 'default', alpha = 10**-3,
-             max_iter = 20000, crit = 1e-1000):
+             max_iter = 200000, crit = 1e-1000):
     '''
     function takes
     - X: n x p dataset
@@ -496,7 +496,7 @@ def scotGradAsc(sigma, V, reg_param, x0 = 'default',
     return v
 
 def SCoT(X, reg_param, x0 = 'default', alpha = 10**-3,
-             max_iter = 20000
+             max_iter = 200000
          
          , crit = 1e-1000):
     '''

### 2020-05-04 Presentation/Presentation.tex

@@
 \usepackage{graphicx}
+\usepackage{apacite}
 \author{
@@
 \title{Sparse Principal Components Analysis}
-%\setbeamercovered{transparent} 
-%\setbeamertemplate{navigation symbols}{} 
-%\logo{} 
+% \setbeamercovered{transparent} 
+% \setbeamertemplate{navigation symbols}{} 
+% \logo{} 
 \institute{\includegraphics{../Figures/nyu_long_black.png}} 
@@
 \begin{itemize}
-\item 
+    \item PCA is a popular dimension reduction tool with applications across many domains; however, one drawback of the technique is that PCs are usually a linear combination of \textit{all} variables and are thus difficult to interpret
+    \begin{itemize}
+        \item This is especially challenging in high-dimension, low sample size (HDLSS) data sets (i.e. "fat" matrices)
+        \item PCA may also lead to inconsistent results in HDLSS data
+    \end{itemize}
+    \item \textbf{Sparse PCA} methods have been developed in an effort to compute more interpretable PCs with few nonzero loadings
+    \item In our project, we implemented three of these sparse PCA methods and evaluated their results on breast tissue data set
 \end{itemize}
@@
 \begin{itemize}
-\item 
+    \item Simple thresholding, rotation and integer programming methods are sometimes used in practice to obtain sparse PCs, but these methods have been shown to produce non-optimal or even misleading results
+    \item Other sparse PCA techniques such as \textbf{SCoT}, \textbf{SCoTLASS} and \textbf{SPCA} have been developed since the early 2000s that more effectively produce sparse PCs that maximize explained variance
+    \begin{itemize}
+    \item Further adaptations and improvements have been made on these sparse PCA techniques, but they continue to be frequently cited in related literature as core methods
+    \end{itemize}
 \end{itemize}
@@
 
-\begin{frame}\frametitle{Methodology}
+\begin{frame}\frametitle{Methodology: Selected Sparse PCA Methods}
+We implemented the \textbf{SCoT}, \textbf{SCoTLASS} and \textbf{SPCA} methods and on a breast tissue measurements data set with n = 106 and p = 9\\
 \begin{itemize}
-\item 
+    \item \textbf{SCoT}: This method maximizes both the regular PCA objective and the \textit{varimax} criterion, which is a sparsity criterion that maximizes the variance of the squared loadings
+    \item \textbf{SCoTLASS}: SCoTLASS induces sparse loadings by maximizing the regular PCA subject to an \textit{L1-norm} constraint
+    \item \textbf{SPCA}: SPCA re-configures the sparse PCA problem into a \textit{regression problem} and uses an \textit{elastic net} constraint to efficiently produce sparse loadings 
+\end{itemize}
+To optimize the SCoT and SCoTLASS objective functions, we used projected gradient ascent
+\begin{itemize}
+    \item Due to the non-differentiable nature of the L1-norm, we approximated the norm with a smooth function (this procedure was outlined by Trendafilov (2004))
+\end{itemize}
+\end{frame}
+
+\begin{frame}\frametitle{Methodology: Evaluation}
+We evaluated these three sparse PCA methods based on the following three criteria:
+\begin{itemize}
+    \item \textit{Sparsity}: We looked at sparsity in terms of the \textit{gini coefficient}, which is a metric for weight distribution across loadings
+        \begin{itemize}
+            \item The Gini coefficient can take on a value between 0 and 1 inclusive, with 0 indicating minimal sparsity (i.e.
+all loadings have equal weight) and 1 indicating maximum sparsity (i.e.
+all weight on one loading)
+        \end{itemize}
+    \item \textit{Explained variance}: We calculated \textit{percentage of explained variance (PEV)} of the sparse PCs that resulted from each method
+    \item \textit{Efficiency and scalability}: For each method, we measured run time on the breast tissue data set assessed the method's suitability for higher dimensional data
 \end{itemize}
@@
 \begin{itemize}
-\item 
+\item SPCA proved to be the most efficient and scalable method that we explored
+\begin{itemize}
+    \item Both SCoT and SCoTLASS require hyperparameter selection, which turned out to be fairly tricky and time consuming; in addition, the non-convex nature of the SCoTLASS objective can result in many local maxima, and thus several random initialization points may be necessary to find the global maximum
+\end{itemize}
 \end{itemize}

### 2020-05-05 Scripts/sparse_pca.py
Updated tables and plots with new EV calculations

@@ -631,3 +631,14 @@ def LARS_EN(Y, X, reg_param, reg_param1):
     beta = lasso.coef_/np.sqrt(1 + reg_param)
     
     return beta
+
+def Adj_Var(X, PCs):
+    adj_EVs = []
+    X_prior = np.zeros((X.shape[1],X.shape[1]))
+    for i in range(0,len(PCs)):
+        V = np.reshape(np.array([PCs[j] for j in range(i+1)]),(i+1,len(PCs[i])))
+        X_k = X @ V.T @ np.linalg.inv(V @ V.T) @ V
+        adj_EV_k = (np.trace(X_k.T@X_k) - np.trace(X_prior.T @ X_prior))/(X.shape[0]-1)
+        adj_EVs.append(adj_EV_k)
+        X_prior = X_k
+    return np.array(adj_EVs)

### 2020-05-06 Scripts/sparse_pca.py
fixed gini function (abs val)

@@ -19,7 +19,7 @@ def gini(array):
     # from:
     # http://www.statsdirect.com/help/default.htm#nonparametric_methods/gini.htm
     # All values are treated equally, arrays must be 1d:
-    array = array.flatten()
+    array = np.abs(array.flatten())
     if np.amin(array) < 0:
         # Values cannot be negative:
         array -= np.amin(array)

### 2020-05-08 Paper/Paper.tex
Updates from Overleaf

@@
 \usepackage{xcolor}
+\usepackage{siunitx}
 \usepackage{upgreek}
 \usepackage{apacite}
+\usepackage{booktabs}
 \usepackage{ntheorem}
@@
 \begin{aligned}
-\arg\max_{v_{i}} \quad & \underbrace{v_{i}^{T}\Sigma v_{i}}_{\text{Variance}} + \lambda \underbrace{S(v_{i})}_{\text{Sparsity}}\\
+\arg\max_{v_{i}} \quad & \underbrace{v_{i}^{T}\Sigma v_{i}}_{\text{Variance}} +\ \theta\underbrace{S(v_{i})}_{\text{Sparsity}}\\
 \textrm{s.t.} \quad & \lVert v_{i} \rVert_{2} = 1 \text{ and } v_{i}^{T}v_{j} = 0 \text{ for } j < i < p \\
@@
 $S(v_i)$ is often referred to as the varimax rotation criteria and simplifies the structure of vector $v_i$ through driving the several loading to zero by maximizing the variance of the squared vector loadings.
-The important of this sparsity components in the maximization problem is regulated by the term $\lambda$.
+The important of this sparsity components in the maximization problem is regulated by the term $\theta$.
 Since both the variance and sparsity terms in the maximization problem are easily differentiable, we chose to implement this method through the use of gradient descent.
@@
 
-\begin{equation}
+\begin{equation*}
 \begin{aligned}
@@
 \end{aligned}
-\end{equation}
+\end{equation*}
 
 Similarly to SCoT, we used projected gradient ascent to optimize SCoTLASS.
-However, the SCoTLASS problem involves an L1 penalty constraint, which is not differentiable at every point and thus poses an issue when using projected gradient ascent.\\
+However, the SCoTLASS problem involves a LASSO constraint, which is not differentiable at every point and thus poses an issue when using gradient ascent.\\
 \\
@@
 This procedure is described in detail by Trendafilov \citeyear{Trendafilov2004}.
-These modifications result in the following adapted SCoTLASS optimization problem for some regularization parameter $t$ and large $\mu$ (we used $\mu = 1000$):
+These modifications result in the following adapted SCoTLASS optimization problem for some regularization parameter $t$ and large $\mu$ (we used $\mu = 1000$) which we solve using projected gradient descent:
 
-\begin{equation}
+\begin{equation*}
 \begin{aligned}
@@
 \end{aligned}
-\end{equation}
+\end{equation*}
 
@@
 \begin{equation*}
-\arg\max_{v_{i}} \quad \|Z_i - X v_i\|^2 + \lambda\|v_i\|^2 + \lambda_1\|v_i\|_1
+\arg\max_{v_{i}} \quad \|Z_i - X v_i\|_2^2 + \lambda_1\|v_i\|_1 + \lambda_2\|v_i\|_2^2
 \end{equation*}
 If we look at this equation we recognize a connection between PCA and the regression method.
-The addition of the positive ridge penalty $\lambda\|v_i\|^2$ ensures that even in the case of a $X$ where is not full-rank because $p>>n$ we will find a unique solution for $v_i$.
+The addition of the positive ridge penalty $\lambda_2\|v_i\|_2^2$ ensures that even in the case of a $X$ where is not full-rank because $p>>n$ we will find a unique solution for $v_i$.
 Zou et al.
@@
 \begin{aligned}
-\arg\max_{B} \quad & \sum_{i=1}^{k} \left(\|X\alpha_j-X\beta_j\|^2 + \lambda\|\beta_j\|^2 +\lambda_{1,j} \|\beta_j\|_1\right)\\
+\arg\max_{B} \quad & \sum_{i=1}^{k} \left(\|X\alpha_j-X\beta_j\|_2^2 +\lambda_1 \|\beta_j\|_1 + \lambda_2\|\beta_j\|_2^2\right)\\
 \textrm{s.t.} \quad & \|\alpha\|^2 = 1\\
@@
 In order to numerically solve this maximization problem we first look for $B$ given $A$ and then $A$ given $B$ in an alternating fashion until $A$ converges.
-The general SPCA algorithm looks like:\vspace*{1mm}\\
-\begin{algorithm}[H]
-\SetAlgoLined
-\KwResult{$V = [v_1,\ldots,v_k]$ where $\|v_j\|=1$}
-$A_0 = Z$\;
-\While{$\|A_{t+1} - A_{t}\|_F \ne 0$}{
-    \For{$\beta_j$ in $B_{t+1}$}{
-    $Y_j = X\alpha_j$\;
-    $\beta_j = \arg\min_\beta \|Y_j - X\beta_j\|^2 + \lambda\|\beta\|^2 + \lambda_{1,j}\|\beta\|_1$\;
-    }
-    $X^TXB_{t+1}=UDV^T$\;
-    $A_{t+1} = UV^T$\;
-}
-\For{$\alpha_j$ in $A$}{
-$v_j = \alpha_j/\|\alpha_j\|$\;
-}
-\caption{General SPCA Algorithm}
-\label{algo1:SPCA}
-\end{algorithm}\vspace*{1mm}
+The general SPCA algorithm looks like algorithm \ref{algo1:SPCA}.\\
+
 In order to solve line $5$ in the algorithm we apply the LARS-EN method for elastic net regressions introduced by Zou and Hastie \cite{Zou2005} which is itself an adaptation of Efron's \citeyear{Efron2004} LARS method.
@@
 
-We evaluated these methods based on three criteria: sparsity, percentage of explained variance (PEV), and efficiency and scalability.\\
+We evaluated these methods based on three criteria: total explained variance, concentration of explained variance, sparsity, and efficiency.\\
+
+\textit{Total Explained Variance:} Most of the method we review in this paper yield orthogonal principal components.
+Thus, they will span the observation space of $X$ explaining all the available variance among themselves.
+However, the SPCA method does enforce orthogonality upon its principal components and hence these sparse components will not explained all the variance in our data set.
+We calculate the total variance explained by the non-orthogonal components as the total variance explained by the orthogonal basis of the span of these non-orthogonal components.
+This orthogonal basis is derived using SVD.\\
+
+\textit{Concentration of Explained Variance:} Since the main objective of PCA is dimensionality reduction, you would want to generate few components that explain the majority of the variance within the dataset.
+Essentially, an unequal distribution of explained variance among the principal components is beneficial to dimensionality reduction.
+Here we use the Gini coefficient introduced by \citeA{Gini1912} as an estimator for the inequality in explanatory value among the principal components.
+The coefficient takes a value between 0 and 1 inclusive, with a 0 value indicating total equality of loadings and a value of 1 indicating total inequality (i.e.
+there is only one nonzero loading).
+Thus, a larger gini coefficient corresponds with a more concentrated explanatory value among the components.\\
 
 \textit{Sparsity:} Since the sparse PCA methods we evaluated often result in nonzero loadings that are very close to zero, we cannot simply count the number of zero loadings as a measure of sparsity.
-For this reason, we used the \textit{gini coefficient} to measure sparsity of loadings as recommended by \citeA{Hurley2008}.
-The gini coefficient is a simple measure of weight distribution across a vector.
-The coefficient takes a value between 0 and 1 inclusive, with a 0 value indicating total equality of loadings and a value of 1 indicating total inequality (i.e.
-there is only one nonzero loading).
-Thus, a larger gini coefficient corresponds with higher sparsity.
-\\
+For this reason, we reused the gini coefficient to measure sparsity of loadings as recommended by \citeA{Hurley2008}.
+An overal sparsity indicator per method is derived by taking the weighted average of the gini coefficients of the componets in regard with the percentage of the explained variance of that component.
+This way we will value those component with a lot of explanatory value more.\\
 
-\textit{Percentage of explained variance (PEV):} An effective sparse PCA method will achieve sparsity while maintaining as much explained variance in the first few principal components as possible.
-For each sparse PCA method, we calculated cumulative PEV (CPEV) of the PCs and plotted the results for comparison purposes.
-\\
-
-\textit{Efficiency and scalability:} Certain sparse PCA methods work well with small data sets with few variables, but become computationally impractical with higher-dimensional data sets.
-For each method, we measured run time on the breast tissue data set and analyzed level of complexity to assess the suitability of each method for different data set types.\\
-
-%Results: What results did you obtain? Do they make sense? Provide a thorough analysis. Negative results are completely fine (they can be very valuable!).
+\textit{Efficiency (Run Time):} Certain sparse PCA methods work well with small data sets with few variables, but become computationally impractical with higher-dimensional data sets.
+For each method, we measured run time on the breast tissue data set and analyzed level of complexity to assess the suitability of each method for different data set types.
 
 \section*{Results}
+%Results: What results did you obtain? Do they make sense? Provide a thorough analysis. Negative results are completely fine (they can be very valuable!).
 
@@
 
+\begin{table}[htpb!]
+    \centering
+    \caption{Variables ordered by loading magnitude (1 = largest loading weight)}
+    \label{rstl:loadings}
+    \input{../Tables/orderedLoads.tex}
+\end{table}
+
 We find that, as we begin to constrain loadings, both SCoT and SCoTLASS shift the majority of weight from variable 4 to variable 7.
-Variable 6 also surprisingly jumps from 6th largest weight in PCA to 3rd largest weight.
+Variable 6 also surprisingly jumps from 6th largest weight in PCA to 3rd largest weight in SCoTLASS and 2nd largest weight .
 This indicates that there may be more predictive significance in these two variables than regular PCs would suggest.
@@
 SCoTLASS is particularly costly to optimize due to the objective function's nonconvexity.
-This results in possible local maxima, therefore requiring several random starting points to obtain the global maxima.
-\\
+This results in possible local maxima, therefore requiring several random starting points to obtain the global maxima.\\
 
@@
 At $\lambda$ = 650 however, the PC 1 that results has six zero loadings, with one of the three nonzero loadings carrying nearly all the weight with a 0.98 coefficient, and explains only 20\% of variance in the data.
-This phenomenon makes SCoT an even more challenging model to tune.
+The sensitivity of the model makes SCoT an even more challenging model to tune.
 
@@
 \section*{Appendix}
+
+\begin{algorithm}[H]
+\SetAlgoLined
+\KwResult{$V = [v_1,\ldots,v_k]$ where $\|v_j\|=1$}
+$A_0 = Z$\;
+\While{$\|A_{t+1} - A_{t}\|_F \ne 0$}{
+    \For{$\beta_j$ in $B_{t+1}$}{
+    $Y_j = X\alpha_j$\;
+    $\beta_j = \arg\min_\beta \|Y_j - X\beta_j\|^2 + \lambda\|\beta\|^2 + \lambda_{1,j}\|\beta\|_1$\;
+    }
+    $X^TXB_{t+1}=UDV^T$\;
+    $A_{t+1} = UV^T$\;
+}
+\For{$\alpha_j$ in $A$}{
+$v_j = \alpha_j/\|\alpha_j\|$\;
+}
+\caption{General SPCA Algorithm}
+\label{algo1:SPCA}
+\end{algorithm}
+
 \begin{table}[htpb!]
@@
     \label{rstl:scot}
-    \include{../Tables/PCA.tex}
+    \input{../Tables/SCoT}
+\end{table}
+
+\begin{table}[htpb!]
+    \centering
+    \caption{SCoTLASS Results for different parameters}
+    \label{rstl:scotlass}
+    \input{../Tables/SCoTLASS}
+\end{table}
+
+\begin{table}[htpb!]
+    \centering
+    \caption{SPCA Results for different parameters}
+    \label{rstl:spca}
+    \input{../Tables/SPCA}
 \end{table}

### 2020-05-08 Paper/Paper.tex
Updates from Overleaf

@@
 \usepackage{ntheorem}
+\usepackage{subcaption}
 \usepackage{enumitem}
@@
 \input{macros}
+
+\makeatletter
+\def\@makechapterhead#1{%
+  %%%%\vspace*{50\p@}% %%% removed!
+  {\parindent \z@ \raggedright \normalfont
+    \ifnum \c@secnumdepth >\m@ne
+        \huge\bfseries \@chapapp\space \thechapter
+        \par\nobreak
+        \vskip 20\p@
+    \fi
+    \interlinepenalty\@M
+    \Huge \bfseries #1\par\nobreak
+    \vskip 40\p@
+  }}
+\def\@makeschapterhead#1{%
+  %%%%%\vspace*{50\p@}% %%% removed!
+  {\parindent \z@ \raggedright
+    \normalfont
+    \interlinepenalty\@M
+    \Huge \bfseries  #1\par\nobreak
+    \vskip 40\p@
+  }}
+\makeatother
 
@@
 
-\textit{Total Explained Variance:} Most of the method we review in this paper yield orthogonal principal components.
-Thus, they will span the observation space of $X$ explaining all the available variance among themselves.
-However, the SPCA method does enforce orthogonality upon its principal components and hence these sparse components will not explained all the variance in our data set.
-We calculate the total variance explained by the non-orthogonal components as the total variance explained by the orthogonal basis of the span of these non-orthogonal components.
-This orthogonal basis is derived using SVD.\\
+\textit{Total Explained Variance:} Most of the method we review in this paper yield orthogonal PCs.
+Thus, summing the explained variance of each resulting PC gives us the total variance of the data.
+However, the SPCA method does not enforce orthogonality upon its PCs and hence summing the explained variance of each resulting PC would be too generous an estimate for total variance.
+Therefore, we calculated \textit{adjusted explained variance} for the each method, as described in \citeyear{Shen2008}, which reflects the incremental increase in explained variance that each additional PC provides (see Appendix for adjusted explained variance algorithm).
+For SCoT and SCoTLASS, adjusted explained variance is the same as regular explained variance.\\
 
@@
 
-The table below gives a summary of the results of the different sparse PCA methods for varying regularization parameters.\\
-
-{\color{red}[TABLE HERE]
-We want four rows: PCA, SCoT, SCoTLASS, SPCA for their "best" specification.
-We will show the difference within each model in tables and figures in the appendix.
-Columns containing: Weighted Gini coefficient, distribution of explained variance, }
-\\
+Table \ref{tab:sumStats} shows a summary of the results of the different sparse PCA methods for varying regularization parameters.\\
 
 \begin{table}[htpb!]
+    \begin{minipage}{.49\linewidth}
+        \caption{Summary Statistics}
+        \label{tab:sumStats}
         \centering
-    \caption{Variables ordered by loading magnitude (1 = largest loading weight)}
-    \label{rstl:loadings}
+        \input{../Tables/SummaryModels.tex}
+    \end{minipage}
+    \begin{minipage}{.49\linewidth}
+        \caption{Sorted Loadings}
+        \label{tab:sorLoad}
+        \centering
         \input{../Tables/orderedLoads.tex}
+    \end{minipage}
 \end{table}
@@
 
+\begin{figure}[htpb!]\centering
+    \caption{Sparsity of Components}
+    \label{fig:sparComp}
+    \includegraphics[scale = 0.5]{"../Figures/sparsityOfComponents.pdf"}
+\end{figure}
+
 \section*{Discussion}
+We notice while PCA delivers us orthogonal principal components with optimal front-loaded explained variance, they are unintelligible.
+Over the years several sparse PCA algorithms have been introduced such as SCoT, SCoTLASS and SPCA.
+We find that all of them make a sparsity versus front-loaded variance trade-off.
+After comparing all models we conclude that SPCA comes out as the first his class with {\color{red} continue...}, but does not yield orthogonal components like the other options.\\
+
+Since SPCA's components are correlated, an interesting future topic of research could be finding a way to transform these into orthogonal set of SPCA components.
+Moreover, while LARS and proximal gradient descent can be used to solve for a LASSO component, a lot of efficiency could be gained by a faster algorithm.
+As we tried working with a dataset of $>20k$ observations, we notice the many limitations these wide dataset enforce on PCA methods and their computability.
 %Discussion: What did you find out? Do the results suggest any other interesting questions to explore?
@@
 
+\begin{figure}
+\begin{subfigure}[htpb!]{0.49\textwidth}
+    \centering
+    \caption{Explained Variance for SCoT}
+    \label{fig:EV_SCoT}
+    \includegraphics[scale = 0.5]{"../Figures/SCoT_comparison.pdf"}
+\end{subfigure}
+\begin{subfigure}[htpb!]{0.49\textwidth}
+    \centering
+    \caption{Explained Variance for SCoTLASS}
+    \label{fig:EV_SCoTLASS}
+    \includegraphics[scale = 0.5]{"../Figures/SCoTLASS_comparison.pdf"}
+\end{subfigure}
+\end{figure}
+
 \end{document}

### 2020-05-08 Scripts/sparse_pca.py
updated gini function

@@ -20,10 +20,10 @@ def gini(array):
     # http://www.statsdirect.com/help/default.htm#nonparametric_methods/gini.htm
     # All values are treated equally, arrays must be 1d:
     array = np.abs(array.flatten())
-    if np.amin(array) < 0:
-        # Values cannot be negative:
-        array -= np.amin(array)
-    # Values cannot be 0:
+#     if np.amin(array) < 0:
+#         # Values cannot be negative:
+#         array -= np.amin(array)
+#     # Values cannot be 0:
     array += 0.0000001
     # Values must be sorted:
     array = np.sort(array)

### 2020-05-08 Scripts/sparse_pca.py
added spca function

@@ -632,6 +632,13 @@ def LARS_EN(Y, X, reg_param, reg_param1):
     
     return beta
 
+def SPCA(X, reg, reg2):
+    X = StandardScaler().fit_transform(X)
+    transformer = SparsePCA(n_components = 9, alpha = reg, ridge_alpha = reg2)
+    transformer.fit(X)
+    norm_comps = np.array([i/np.linalg.norm(i) for i in transformer.components_])
+    return norm_comps
+
 def Adj_Var(X, PCs):
     adj_EVs = []
     X_prior = np.zeros((X.shape[1],X.shape[1]))

### 2020-05-08 Scripts/sparse_pca.py
updates

@@ -8,6 +8,7 @@ from scipy.optimize import minimize
 from scipy.linalg import null_space
 from IPython.display import clear_output
 from sklearn.linear_model import LassoLars
+from sklearn.decomposition import SparsePCA
 import random
 import warnings
 warnings.filterwarnings('ignore')

### 2020-05-10 Scripts/sparse_pca.py
edits to plots and tables

@@ -282,7 +282,7 @@ def scotlass_grad(sigma, v, reg_param, gamma):
     return (sigma @ v) - 0.5*mu*q*z 
 
 def scotlassGradAsc(sigma, V, reg_param , x0 = 'default', 
-                    alpha = 10**-3, max_iter = 20000, crit = 1e-1000):
+                    alpha = 10**-6, max_iter = 500000, crit = 1e-1000):
     '''
     function takes
     - sigma: p x p covariance matrix
@@ -342,8 +342,8 @@ def scotlassGradAsc(sigma, V, reg_param , x0 = 'default',
     # return loadings array v
     return v
 
-def SCoTLASS(X, reg_param, x0 = 'default', alpha = 10**-3,
-             max_iter = 200000, crit = 1e-1000):
+def SCoTLASS(X, reg_param, x0 = 'default', alpha = 10**-6,
+             max_iter = 500000, crit = 1e-1000):
     '''
     function takes
     - X: n x p dataset
@@ -444,7 +444,7 @@ def scot_obj_grad(sigma, v, reg_param):
     return grad
 
 def scotGradAsc(sigma, V, reg_param, x0 = 'default', 
-                    alpha = 10**-3, max_iter = 20000, crit = 1e-1000):
+                    alpha = 10**-5, max_iter = 500000, crit = 1e-1000):
     '''
     function takes
     - sigma: p x p covariance matrix
@@ -496,10 +496,8 @@ def scotGradAsc(sigma, V, reg_param, x0 = 'default',
     # return loadings array v
     return v
 
-def SCoT(X, reg_param, x0 = 'default', alpha = 10**-3,
-             max_iter = 200000
-         
-         , crit = 1e-1000):
+def SCoT(X, reg_param, x0 = 'default', alpha = 10**-5,
+             max_iter = 500000, crit = 1e-1000):
     '''
     function takes
     - X: n x p dataset

### 2020-05-10 Paper/Paper.tex
Updates from Overleaf

@@
 PCA does this by computing orthogonal linear combinations of the original variables, called the principal components (PCs), such that the variance along these components are maximized.
-The resulting PCs are ordered from highest variance to lowest variance, and thus the first few principal components will retain the most variance in the data.\\
+The resulting PCs are ordered from highest variance to lowest variance, and thus the first few PCs will retain the most variance in the data.\\
 \\
@@
 This is particularly challenging when the number of variables $p$ is much larger than the number of examples $n$ (commonly called “High Dimension, Low Sample Size” data, or HDLSS), which is the case for many modern data sets.
-Consequently, much effort has gone into developing methods of approximating PCs with fewer than the $p$ original variables.\\
+Consequently, much effort has gone into developing methods of approximating PCs dependent on fewer than the $p$ original variables.\\
 \\
 In this paper, we will explore several sparse PCA methods that have been developed over the past few decades and discuss the benefits and drawbacks of each method.
-We will evaluate each method on a breast tissue data set compiled by JP Marques de Sá at INEB-Instituto de Engenharia Biomédica.
+We will evaluate each method on a breast tissue data set analyzed by Marques de Sá and Jossinet \citeyear{da2000}.
 
@@
 It has since been a staple in statistics curricula and academic research.
-As previously mentioned, one of the most relevant drawbacks to using PCA is the difficulty of interpreting separate principal components as they are a linear combination of all the original variables.
-Since its creation, several attempts have been made at making the principal components more intelligible.\\
+As previously mentioned, one of the most relevant drawbacks to using PCA is the difficulty of interpreting separate PCs as they are a linear combination of all the original variables.
+Since its creation, several attempts have been made at making the PCs more intelligible.\\
 \\
@@
 One of those approaches frequently applied is setting loadings with absolute values below a certain threshold to zero.
-This \textit{thresholding} method could lead to misleading results for a series of reasons as described by Cadima and Jolliffe \citeyear{Cadima1995}.
+Other simple sparse PCA methods include performing orthogonal rotations on the regular PCA components to maximize simplicity criteria such as \textit{varimax} and \textit{quartimax} \cite{Richman1986, Richman1987} and using integer programming to restrict loadings to be from a finite set of integers \cite{Hausman1982, Vines2000}.
+Though useful in certain situations, these methods have many drawbacks and can produce misleading results.\\
+
 % Orthogonal rotations
-Another common practice based on factor  analysis is to perform and orthogonal rotation on the regular PCA components \cite{Richman1986, Richman1987}.
+% Another common practice based on factor  analysis is to perform and orthogonal rotation on the regular PCA components \cite{Richman1986, Richman1987}.
 This rotation is then optimized by maximizing simplicity criteria such as \textit{varimax} and \textit{quartimax}.
@@
 The drawback of this method is that it does not preserve the successive maximization property inherent to PCA.
-Moreover Jolliffe \citeyear{Jolliffe1989, Jolliffe1995} warns that orthogonal rotations often lead to non-orthogonal loadings and pairwise correlation between the principal components, and should be reserved for principal components with near-equal variance.\\
+Moreover Jolliffe \citeyear{Jolliffe1989, Jolliffe1995} warns that orthogonal rotations often lead to non-orthogonal loadings and pairwise correlation between the PCs, and should be reserved for PCs with near-equal variance.\\
+
+% SCoT
+In 2000, \citeauthor{Jolliffe2000} introduced the \textit{simple component technique (SCoT)} approach which combines the objectives of variance maximization (like PCA) and simplification (using the varimax criteria).
+Contrary to a the normal orthogonal rotation approach that stays within the subspace with maximum variance defined by the regular principal directions, SCoT's use of a penalty function will make the resulting PCs divert from this maximum variance subspace and hence faces a variance-simplicity trade-off.\\
 \\
-% SCoT
-This rotation method is two-staged in nature as it requires you to first compute PCA and then rotate the resulting PCs optimally.
-However, Jolliffe and Uddin \citeyear{Jolliffe2000} introduced the one-stage \textit{SCoT} approach that combines the objectives of variance maximization (like PCA) and simplification (using the varimax criteria).
-Contrary to a the normal orthogonal rotation approach that stays within the subspace with maximum variance defined by the regular principal directions, SCoT's use of a penalty function will make the resulting principal components divert from this maximum variance subspace and hence faces a variance-simplicity trade-off.\\
-\\
-% Vines
-While SCoT yields increased sparsity compared to PCA, the loadings will still be nonzero values.
+% % Vines
+% While SCoT yields increased sparsity compared to PCA, the loadings will still be nonzero values.
 This is an issue that both Hausman \citeyear{Hausman1982} and Vines \citeyear{Vines2000} aim to solve.
-Hausman introduced an integer programming solution for assembling principal components using integers loadings from a finite set (typically {$-1$,$0$,$1$}).
+Hausman introduced an integer programming solution for assembling PCs using integers loadings from a finite set (typically {$-1$,$0$,$1$}).
 Each element is determined sequentially as to optimize the variance explained by each component.\\
-\\
+% \\
 % SCoTLASS
@@
 % SPCA
-One of the later developments is sparse PCA (henceforth SPCA) developed by Zou, Hastie and Tibshirani \citeyear{Zou2006}.
+One of the later developments is \textit{sparse PCA (SPCA)} developed by Zou, Hastie and Tibshirani \citeyear{Zou2006}.
 Like SCoTLASS, this method incorporates the lasso method, but in addition, they reformulate the PCA in a regression-type optimization problem.
-This method achieves overall sparse loadings with limited variable overlap between the different principal components.
-However, unlike SCoT and SCoTLASS, SPCA results in non-orthogonal PCs.
-In the years following its first publication, SPCA has been improved upon in a plethora of ways \cite{Shen2008, Journee2010}.\\
-\\
+This method achieves overall sparse loadings with limited variable overlap between the different PCs.
+However, unlike SCoT and SCoTLASS, SPCA results in non-orthogonal PCs.\\
+
 % After SPCA
-Developments in the field of sparse PCA after the introduction of SPCA mainly build on the spiked covariance model introduced by Johnstone \citeyear{Johnstone2001}.
+% Developments in the field of sparse PCA after the introduction of SPCA mainly build on the spiked covariance model introduced by Johnstone \citeyear{Johnstone2001}.
 The model assumes that there exist a small number of natural directions that explains most of the variance.
@@
 Further research focuses on solving this computational problem\footnote{Such as \citeA{Wang2016, Cai2013, Vu2013, Berthet2013}.}.
+
+Many adaptations and improvements have been made to these models since their conception. Some of these methods are outlined in \citeA{Shen2008, Wang2016, Cai2013, Vu2013, Berthet2013}
 
@@
 $S(v_i)$ is often referred to as the varimax rotation criteria and simplifies the structure of vector $v_i$ through driving the several loading to zero by maximizing the variance of the squared vector loadings.
-The important of this sparsity components in the maximization problem is regulated by the term $\theta$.
+The important of this sparsity component in the maximization problem is regulated by the term $\theta$.
 Since both the variance and sparsity terms in the maximization problem are easily differentiable, we chose to implement this method through the use of gradient descent.
@@
 
-\subsection*{Sparse Principal Components (SPCA)}
+\subsection*{Sparse PCs (SPCA)}
 SPCA was introduced by Zou et al.
-\cite{Zou2006} and applies the elastic net to find principal components with sparse loadings.
-They start form the regular PCA principal components of the zero-mean dataset $X$ with as SVD decomposition $UDV^T$ where $Z=UD$ are the uncorrelated and orthogonal principal components.
+\cite{Zou2006} and applies the elastic net to find PCs with sparse loadings.
+They start form the regular PCA PCs of the zero-mean dataset $X$ with as SVD decomposition $UDV^T$ where $Z=UD$ are the uncorrelated and orthogonal PCs.
 SPCA will attempt to find a $v_i$ that is a sparse estimate of the $i$th principal component $Z_i = U_iD_{ii}$ regularized with an elastic net.
@@
 Zou et al.
-continue with this similarity between PCA and a elastic net regression and derive the following optimization problem for the $k$ first principal components:
+continue with this similarity between PCA and a elastic net regression and derive the following optimization problem for the $k$ first PCs:
 \begin{equation*}
@@
 
-In order to solve line $5$ in the algorithm we apply the LARS-EN method for elastic net regressions introduced by Zou and Hastie \cite{Zou2005} which is itself an adaptation of Efron's \citeyear{Efron2004} LARS method.
+In order to solve line $5$ in the algorithm \ref{algo1:SPCA} we apply the LARS-EN method for elastic net regressions introduced by Zou and Hastie \cite{Zou2005} which is itself an adaptation of Efron's \citeyear{Efron2004} LARS method.
 Essentially, we are able to transform an elastic net problem to a simple lasso problem and apply LARS by constructing an artificial data matrices $X^*$ and $Y^*$ based on $X$, $Y$ and $\lambda$.
@@
 The data set had 9 variables and 106 examples (see Table \ref{tab1:dta_vars} for descriptions of each variable).
-\\
 
@@
 
-We evaluated these methods based on three criteria: total explained variance, concentration of explained variance, sparsity, and efficiency.\\
+We evaluated these methods based on three criteria: sparsity, explained variance and efficiency.\\
 
-\textit{Total Explained Variance:} Most of the method we review in this paper yield orthogonal PCs.
-Thus, summing the explained variance of each resulting PC gives us the total variance of the data.
-However, the SPCA method does not enforce orthogonality upon its PCs and hence summing the explained variance of each resulting PC would be too generous an estimate for total variance.
-Therefore, we calculated \textit{adjusted explained variance} for the each method, as described in \citeyear{Shen2008}, which reflects the incremental increase in explained variance that each additional PC provides (see Appendix for adjusted explained variance algorithm).
-For SCoT and SCoTLASS, adjusted explained variance is the same as regular explained variance.\\
-
-\textit{Concentration of Explained Variance:} Since the main objective of PCA is dimensionality reduction, you would want to generate few components that explain the majority of the variance within the dataset.
-Essentially, an unequal distribution of explained variance among the principal components is beneficial to dimensionality reduction.
-Here we use the Gini coefficient introduced by \citeA{Gini1912} as an estimator for the inequality in explanatory value among the principal components.
+\textit{Sparsity:} Since the sparse PCA methods we evaluated often result in nonzero loadings that are very close to zero, we cannot simply count the number of zero loadings as a measure of sparsity.
+For this reason, we used the Gini coefficient introduced by \citeA{Gini1912} to measure sparsity of loadings as recommended by \citeA{Hurley2008}.
 The coefficient takes a value between 0 and 1 inclusive, with a 0 value indicating total equality of loadings and a value of 1 indicating total inequality (i.e.
 there is only one nonzero loading).
-Thus, a larger gini coefficient corresponds with a more concentrated explanatory value among the components.\\
+Thus, a larger Gini coefficient corresponds with more sparsity.\\
 
-\textit{Sparsity:} Since the sparse PCA methods we evaluated often result in nonzero loadings that are very close to zero, we cannot simply count the number of zero loadings as a measure of sparsity.
-For this reason, we reused the gini coefficient to measure sparsity of loadings as recommended by \citeA{Hurley2008}.
-An overal sparsity indicator per method is derived by taking the weighted average of the gini coefficients of the componets in regard with the percentage of the explained variance of that component.
-This way we will value those component with a lot of explanatory value more.\\
+An overall sparsity indicator for each method is derived by taking the weighted average of the Gini coefficients of the PCs, with each Gini coefficient weighted by the PEV of that component.
+Weighting the Gini coefficients by PEV effectively places higher value on those components with higher explanatory power.
+Note that Gini coefficient is only defined for vectors with non-negative loadings, so we calculated the Gini coefficients based on the absolute values of loading vectors.\\
+
+\textit{Explained variance:} SCoT and SCoTLASS methods both yield orthogonal PCs.
+Thus, summing the explained variance (EV) of each resulting PC gives us the total variance of the data.
+However, the SPCA method does not enforce orthogonality upon its PCs, hence summing the EV of each resulting PC would be too generous an estimate for total EV.
+Therefore, for SPCA, we substituted \textit{adjusted explained variance (AEV)} for EV, as described by \citeA{Shen2008}.
+AEV reflects the incremental increase in EV that each additional PC provides.\\
+
+Since the main objective of PCA is dimensionality reduction, our goal is to generate few PCs that explain the majority of the variance within the dataset.
+Essentially, unequal distribution of EV among the PCs is beneficial to dimensionality reduction.
+We calculated the Gini coefficient of EV distribution of resulting PCs (we call this \textit{concentration of EV (CEV)}) for different models to measure inequality in explanatory value among the PCs.
+We also directly compare \textit{percentage of EV (PEV)} of individual PCs.\\
 
 \textit{Efficiency (Run Time):} Certain sparse PCA methods work well with small data sets with few variables, but become computationally impractical with higher-dimensional data sets.
-For each method, we measured run time on the breast tissue data set and analyzed level of complexity to assess the suitability of each method for different data set types.
+For each method, we measured run time on the breast tissue data set.
+%and analyzed level of complexity to assess the suitability of each method for different data set types.
 
@@
 
-Table \ref{tab:sumStats} shows a summary of the results of the different sparse PCA methods for varying regularization parameters.\\
+% \begin{table}[htpb!]
+%     \begin{minipage}{.49\linewidth}
+%         \caption{Summary Statistics}
+%         \label{tab:sumStats}
+%         \centering
+%         \input{../Tables/SummaryModels.tex}
+%     \end{minipage}
+%     \begin{minipage}{.49\linewidth}
+%         \caption{Sorted Loadings}
+%         \label{tab:sorLoad}
+%         \centering
+%         \input{../Tables/orderedLoads.tex}
+%     \end{minipage}
+% \end{table}
 
-\begin{table}[htpb!]
-    \begin{minipage}{.49\linewidth}
-        \caption{Summary Statistics}
-        \label{tab:sumStats}
+\begin{table}[ht]
+\centering
+\caption{Sorted loadings for first PC}\label{tab:sorLoad}
+\begin{tabular}%[htpb!]
+    %\label{tab:sorLoad}
+    \centering
+    \input{../Tables/orderedLoads.tex}
+\end{tabular}
+\vspace*{3mm}
+\caption{Summary statistics}\label{tab:sumStats}
+\begin{tabular}%[htpb!]
+    %\label{tab:sumStats}
     \centering
     \input{../Tables/SummaryModels.tex}
-    \end{minipage}
-    \begin{minipage}{.49\linewidth}
-        \caption{Sorted Loadings}
-        \label{tab:sorLoad}
-        \centering
-        \input{../Tables/orderedLoads.tex}
-    \end{minipage}
+\end{tabular}
 \end{table}
 
-We find that, as we begin to constrain loadings, both SCoT and SCoTLASS shift the majority of weight from variable 4 to variable 7.
-Variable 6 also surprisingly jumps from 6th largest weight in PCA to 3rd largest weight in SCoTLASS and 2nd largest weight .
-This indicates that there may be more predictive significance in these two variables than regular PCs would suggest.
-\\
+Table \ref{tab:sorLoad} indicates that as we begin to constrain loadings, all three methods place heavier importance on variable 7 than on variable 4, which is different from what regular PCA indicates.
+We also see that SCoT and SCoTLASS follow fairly similar loading patterns as we increase regularization strength.
+These two methods also have more consistent changes in loading weights as regularization increases (i.e.
+variables monotonically increase or decrease in importance).
+SPCA loadings differ from SCoT and SCoTLASS loadings, placing most weight on variable 9 as regularization increases.
+It is also more erratic in how it changes weights between variables as regularization increases.
+For example, we can see that variable 9 first falls in importance, then increases to highest importance as regularization increases.\\
 
-In terms of efficiency and runtime, the SPCA function takes considerably less time to run than SCoT and SCoTLASS.
+In terms of sparsity of the first PC, SCoTLASS and SPCA perform similarly to one another.
+We can see that SCoTLASS is able to maintain 0.36 PEV while obtaining a Gini coefficient of 0.60.
+SPCA is able to maintain 0.37 PEV while obtaining a Gini coefficient of 0.61.
+SCoT performed worse than the other two models, only acheiving a Gini coefficient of 0.47 at 0.37 PEV.\\
+
+Table \ref{tab:sumStats} illustrates CEV and weighted average sparsity across all the resulting PCs for each model.
+We see that SCoTLASS has the highest CEV while achieving lowest sparsity.
+SCoT and SPCA perform similarly both in terms of CEV and sparsity across all PCs.
+However, if we are strictly comparing sparsity levels of resulting PCs, Figure \ref{fig:sparComp} below shows that SCoT obtains the most sparse results on average, and SCoTLASS obtains the least sparse results on average.
+In the same figure, we can also see that SCoT and SCoTLASS induce more consistent levels of sparsity among the resulting PCs than SPCA.\\
+
+Table \ref{tab:sumStats} also shows that the SPCA function takes considerably less time to run than SCoT and SCoTLASS.
 This is largely due to our use of projected gradient ascent in maximizing SCoT and SCoTLASS, which required very small step sizes over many iterations in order to for the algorithms to converge.
@@
 
-In optimizing SCoT and SCoTLASS, we found is quite challenging to select the best regularization parameter to yield optimal levels of sparsity and explained variance.
-There is little guidance from relevant literature on selecting regularization parameters for these methods.\\
+In optimizing SCoT and SCoTLASS, we also found it quite challenging to select the best regularization parameter to yield optimal levels of sparsity and explained variance.
+There is little guidance from relevant literature on selecting regularization parameters for these methods.
 
-In addition, as we varied the regularization parameter for SCoT, we also find that there is a sudden increase in loading sparsity, which is accompanied by a sharp decrease in explained variance.
-At $\lambda$ = 600, SCoT produces a PC 1 with one zero loading and most other loadings between 0.30 and 0.44 and explains approximately 60\% of variance in the data.
+% In addition, as we varied the regularization parameter for SCoT, we also find that there is a sudden increase in loading sparsity, which is accompanied by a sharp decrease in explained variance.
+At $\lambda$ = 600, SCoT produces a PC 1 with two  near-zero loading and most other loadings between 0.30 and 0.44 and explains approximately 60\% of variance in the data.
 At $\lambda$ = 650 however, the PC 1 that results has six zero loadings, with one of the three nonzero loadings carrying nearly all the weight with a 0.98 coefficient, and explains only 20\% of variance in the data.
@@
 
-\begin{figure}[htpb!]\centering
-    \caption{Sparsity of Components}
-    \label{fig:sparComp}
-    \includegraphics[scale = 0.5]{"../Figures/sparsityOfComponents.pdf"}
-\end{figure}
-
 \section*{Discussion}
-We notice while PCA delivers us orthogonal principal components with optimal front-loaded explained variance, they are unintelligible.
-Over the years several sparse PCA algorithms have been introduced such as SCoT, SCoTLASS and SPCA.
+While PCA delivers us orthogonal PCs with optimal front-loaded explained variance, they are often unintelligible.
+Over the years several sparse PCA algorithms such as SCoT, SCoTLASS and SPCA have been introduced and improved, particularly for use in high-dimensional datasets.
 We find that all of them make a sparsity versus front-loaded variance trade-off.
-After comparing all models we conclude that SPCA comes out as the first his class with {\color{red} continue...}, but does not yield orthogonal components like the other options.\\
+After comparing all models we conclude that SPCA comes out as the best of class, due to the high level of sparsity that it achieves and its efficiency/scalability, but does not yield orthogonal components like the other options.\\
 
@@
     \centering
-    \caption{SCoT Results for different parameters}
+    \caption{SCoT results for different parameters}
     \label{rstl:scot}
@@
     \centering
-    \caption{SCoTLASS Results for different parameters}
+    \caption{SCoTLASS results for different parameters}
     \label{rstl:scotlass}
@@
     \centering
-    \caption{SPCA Results for different parameters}
+    \caption{SPCA results for different parameters}
     \label{rstl:spca}
@@
 
+\begin{table}[htpb!]
+    \centering
+    \caption{Loadings for first PC}
+    \label{rstl:scot}
+    \input{../Tables/LoadingsPC1}
+\end{table}
+
 \begin{figure}
-\begin{subfigure}[htpb!]{0.49\textwidth}
+\caption{Cumulative Explained Variance per method}
+\begin{subfigure}[H!]{0.49\textwidth}
     \centering
-    \caption{Explained Variance for SCoT}
+    \caption{SCoT}
     \label{fig:EV_SCoT}
@@
 \end{subfigure}
-\begin{subfigure}[htpb!]{0.49\textwidth}
+\begin{subfigure}[H!]{0.49\textwidth}
     \centering
-    \caption{Explained Variance for SCoTLASS}
+    \caption{SCoTLASS}
     \label{fig:EV_SCoTLASS}
@@
 \end{subfigure}
+\begin{center}
+\begin{subfigure}[H!]{0.49\textwidth}
+    \centering
+    \caption{SPCA}
+    \label{fig:EV_SCoTLASS}
+    \includegraphics[scale = 0.5]{"../Figures/SPCA_comparison.pdf"}
+\end{subfigure}    
+\end{center}
+\end{figure}
+
+\begin{figure}[htpb!]\centering
+    \caption{Sparsity of components (ranked from highest to lowest PEV)}
+    \label{fig:sparComp}
+    \includegraphics[scale = 0.5]{"../Figures/sparsityOfComponents.pdf"}
 \end{figure}

### 2020-05-10 Presentation/Presentation.tex
Updates from Overleaf

@@
         \end{itemize}
-    \item \textit{Explained variance}: We calculated \textit{percentage of explained variance (PEV)} of the sparse PCs that resulted from each method
+    \item \textit{Explained variance}: We calculated \textit{percentage of explained variance (PEV)} of the sparse PCs that resulted from each method, as well as \textit{concentration of explained variance (CEV)}
     \item \textit{Efficiency and scalability}: For each method, we measured run time on the breast tissue data set assessed the method's suitability for higher dimensional data
@@
 \begin{itemize}
+\item SCoT and SCoTLASS yielded more similar PCs than SPCA in terms of the ranking of variable importance
+\begin{itemize}
+    \item At higher regularization, SCoT and SCoTLASS selected variable 7 as most important, while SPCA selected variable 9
+\end{itemize}
 \item SPCA proved to be the most efficient and scalable method that we explored
@@
 
-\begin{frame}\frametitle{Discussion}
-\begin{itemize}
-\item 
-\end{itemize}
-\end{frame}
-
 \end{document}
